---
  title: "Semi-targeted Metabolomics Analysis using tsatsiki_v05"
author:
  - name: Dimitrios Prymidis
affiliation:
  - Cologne Universiy
date: 2025-08-20
output: 
  html_document:
  self_contained: true
toc: true
toc_float: true
toc_depth: 5
code_folding: show
vignette: >
  %\VignetteIndexEntry{Frezza lab Metabolomics Analysis Pipeline R script}
%\VignetteEngine{knitr::rmarkdown}
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
---
  
```{r chunk_setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


### Instructions!

# Press Alt+o, and then alt+shift+o , this should minimize your code. Press on the <-> to open specific parts. Now open 0. Settings.

### --- 0. Settings
# Folders
In order to run the pipeline, the folder "tsatsiki_vXX" has to be inside the ProjectFolder. The ProjectFolder has to have the name set as 'Year_name_ProjectCode'.

To Load the data you have to create a folder called "Inputs" in the project folder. In that folder there must be 3 and only files.
1. The SSF, which must be an excel file with the 'SSF' on the file name and it has to have exactly the same structure and the AG.Frezza metabolomics SSF.
2. The data must be an excel file without the 'SSF' in its name. The data must follow the Tracefinder output format (samples in rows and metabolites in columns). 
3. The sample run order in the mass spec. This must be the only csv in the folder. It is found in the Raw data folder. The data should be a table with the names of the samples ordered in the first column.

you will need to add input in
1- 0.2. Set General Parameters: To set up the project folder and folders in general, Thresholds and identifiers.
2- 0.4. Load Data:  This is to fix the sample names in the data to match the SSF and also the names in the runOrder.

Run each chunk by pressing the the Run Current Chunk on the far right side of the code ------------------------> down here!
  
#0.1. Load libraries.
```{r Load libraries}
library(tidyverse)
library(ggfortify)
library(patchwork)
library(reshape2) # for melting
library(ggrepel)  # For smart label positioning
library(gridExtra) # to combine plots
library(ggstatsplot) # for boxplots
library(pheatmap) # Heatmap
library(EnhancedVolcano) # Volcano 
library(MetaboAnalystR) # Ora
library(pathview) # for pathway visualisations
library(openxlsx) # for TF_errors, excel coloring
```

#0.2. Set Parameters.
```{r}
# Instructions

# In order to run the pipeline, the folder "tsatsiki_vXX" has to be inside the ProjectFolder. The ProjectFolder has to have the name set as 'Year_name_ProjectCode'.
# Find the project folder in the windows explorer, copy the file path and run the next line. Then, print and save the path in the variable ProjectFolder. )


### ------------- Set working directory here ----------- ###

# convert_clip <- function() {gsub("\\\\", "/", readClipboard())}
# ProjectFolder <- convert_clip()
# ProjectFolder
ProjectFolder <- "C:/Users/Dimitrios/Desktop/2025_Sebastian_CoRe"

# Get project Code
ProjectCode <- strsplit(ProjectFolder, "_")[[1]][[ length(strsplit(ProjectFolder, "_")[[1]])]]

# Set working directory
setwd(ProjectFolder) 

# Set input folder
input.dir <-  file.path(ProjectFolder, "Inputs")

# Set Output folders
output.dir <-  file.path(ProjectFolder,paste(ProjectCode,"Outputs", sep = " "))

# CoRe Analysis
CoRe = TRUE

### ---------- Set Custom Thresholds ---------- ###

#Metabolite S to N ratio
SignaltoNoiseRatio_threshold = 3

# Feature NA filtering
Feature_Filt_Value <- 0.8 # in 10 samples we allow for 2 NAs 
Feature_Filt_Value_Pools <- 0.8 # If we have 4 pools, we allow 1 NA in 4 which means 75 not 80

# RSd / CV filtering
Threshold_cv = 30

#Filter outlier values - sd threshold ?

# Set Blank sample rownames identifier. This is used to identify the blank samples from the rownames in the data.
blank_samp_id <- "BLK|Blank|blank"

# Set up valine-d8 idenrtifier
valine_pattern_id = "aline-d8|valine-D8"

paste0("#0.2. Set General Parameters Finished!")
```

#0.3. Set universe.
```{r}
Input_name_vec <- c("1,2,3-Propanetricarboxylic acid","1,5-Anhydrosorbitol","1-Methyl nicotinamide","1-Methylhistidine","1-Methylxanthine","2',3' cyclic CMP","2-Aminobenzoic acid","2-Deoxyribose","2-Hydroxyadipic acid","2-Hydroxybutyric acid","2-Hydroxyglutaric acid","2-Hydroxypyridine","2-Oxoadipic acid","2-Oxoglutaric acid","2-Phosphoglyceric acid","3-(3-Hydroxyphenyl)-3-hydroxypropanoic acid","3',5' cyclic AMP","3',5' cyclic GMP","3-Dehydroxycarnitine","3'-Dephospho-CoA","3-Guanidinopropanoic acid","3-Hydroxy-adipic acid","3-Hydroxybutyric acid","3-Hydroxyhippuric acid","3-Hydroxyisobutyric acid","3-Hydroxymethylglutaric acid","3-Methyl pyruvic acid","3-Methylhistamine","3-Methylhistidine","4,5-Dihydroorotic acid","4a-Carbinolamine tetrahydrobiopterin","4-Hydroxyphenyllactic acid","4-Hydroxyphenylpyruvic acid","5-Amino-levulinic acid","5-Hydroxyindoleacetic acid","5-Methylcytosine","5-Phosphoribosylamine","7-Methylguanine","Acetyl-CoA","Adenine","Adenosine","Adenylsuccinic acid","ADP","ADP-glucose","ADP-ribose","AICAR","AIR","Ala-Gln","Ala-Glu","Alanine","Ala-Thr","Ala-Tyr","Allantoin","Alliin","Aminoadipic acid","AMP","Anserine","Arginine","Argininosuccinic acid","Asn-Lys","Asparagine","Aspartic acid","ATP","beta-Alanine","Betaine","Bilirubin","Biliverdin","Biotin","CAR 2:0","CAR 3:0","CAR 3:0;2Me","CAR 4:0;3Me","CAR 4:0;3OH","CAR 4:1;2Me","CAR 5:0","CAR DC3:0","CAR DC4:0","CAR DC5:0","CAR DC6:0","Carbamoylaspartate","Carnitine","Carnosine","CDP","CDP-choline","CDP-Ethanolamine","Choline","Choline phosphate","Citric acid","Citrulline","CMP","CMP-N-acetylneuraminic acid","Coenzyme A","Creatine","Creatinine","CTP","Cyclic ADP-ribose","cyclic CMP","Cys-Gly","Cystathionine","Cysteic acid","Cysteine-S-sulfate","Cysteinylglycine disulfide","Cystine","Cytidine","Cytosine","dAMP","dCMP","Deoxyadenosine","Deoxycytidine","Deoxyguanosine","Deoxyinosine","Deoxyribose 5-phosphate","Deoxyuridine","dGMP","Diethanolamine","Dihydrofolic acid","Dihydrothymine","Dihydrouracil","Dihydroxyacetone phosphate","Dimethylarginine","Dimethylglycine","Dopamine","dTDP","dTDP-glucose","dTMP","dTTP") 
Input_name_vec <- c(Input_name_vec, "dUMP","dUTP","Ectoine","Epsilon-(gamma-Glutamyl)-lysine","Ergothioneine","Erythronic acid","FAD","FAICAR","FGAM","FGAR","FMN","Folic acid","Formiminoglutamic acid","Fructose 6-phosphate","Fumaric acid","Galacturonic acid","gamma-Glutamylcysteine","gamma-Glutamylglutamic acid","gamma-Glutamylglutamine","gamma-Glutamylhistidine","gamma-Glutamylvaline","GAR","GDP","GDP-L-fucose","Glucaric acid","Glucuronic acid","Glutamic acid","Glutamine","Glutarylglycine","Glutathione oxidized","Glutathione","Gly-Asp","Glyceraldehyde 3-phosphate","Glyceric acid","Glycine","Gly-Ile","Gly-Pro","Gly-Tyr","GMP","GTP","Guanidoacetic acid","Guanine","Guanosine","Heme","Hexose 6-phosphate","Histamine","Histidine","His-Val","Homoarginine","Homocitrulline","Homocysteine","Homoserine","Hydroxykynurenine","4-Hydroxyproline","Hypotaurine","Hypoxanthine","Imidazoleacetic acid","IMP","Indoleacetic acid","Indoxyl sulfate","Inosine","Inositol","Isoleucine","Itaconic acid","Ketoisovaleric acid","Ketoleucine","Kynurenic acid","Kynurenine","Lactic acid","Leucine","Lysine","Lys-Val","Maleic acid","Malic acid","Melatonin","Methionine","Methionine sulfone","Methionine sulfoxide","Methylguanidine","Methylimidazoleacetic acid","Methylmalonic acid","Methylthioadenosine","N(2)-Acetyllysine","N(6)-Methyllysine","N1,N12-Diacetylspermine","N1,N8-Diacetylspermidine","N1-Acetylspermine","N6-Acetyllysine","N-6-Trimethyllysine","N8-Acetylspermidine","N-Acetylalanine","N-Acetylasparagine","N-Acetylaspartic acid","N-Acetyl-Asp-Glu","N-Acetylglutamic acid","N-Acetylglutamine","N-Acetylhistidine","N-Acetylneuraminic acid","N-Acetylornithine","N-Acetylserotonin","N-Acetylthreonine","N-Acetyltryptophan","NAD","NADH","NADP","NADPH","N-Carbamoylaspartic acid","N-Formylglycine","N-Formylphenylalanine","Niacinamide","Nicotinamide ribotide","Nicotinic acid","N-Methylserotonin","N-Methyltryptamine","O-Acetylserine","Ophthalmic acid","Ornithine","Orotic acid","Orotidine","o-Succinylhomoserine","Oxoproline","Pantothenic acid","Phenylalanine","Phosphocreatine","Phosphoenolpyruvic acid","Phosphoserine","Picolinic acid","Pipecolic acid","Porphobilinogen","Proline","Proline betaine","Protoporphyrinogen IX","Pseudouridine","Pterin","Purine","Pyridoxal","Pyridoxine","Pyroglutamic acid","Pyruvic acid","Quinic acid","Quinolinic acid","Riboflavin","Ribonic acid","Ribose 5-phosphate","Ribothymidine","Saccharopine","S-Adenosylhomocysteine","S-Adenosylmethionine","SAICAR","Sarcosine","Sedoheptulose 7-phosphate","Ser-Ala","Serine","Serotonin","sn-Glycero-3-phosphate","Sorbitol","Spermidine","Spermine","S-Succinylglutathione","Succinic acid","Succinyl-CoA","Targinine","Taurine","Thiamine","Threonic acid","Threonine","Thymidine","Thymine","trans-Aconitic acid","Trigonelline","Trimethylamine N-oxide","Tryptamine","Tryptophan","Tyrosine","UDP","UDP-glucose","UDP-glucuronic acid","UDP-N-acetylglucosamine","UMP","Uracil","Urea","Ureidopropionic acid","Uric acid","Uridine","Urocanic acid","UTP","Val-Asp","Val-Glu","Valine","Vitamin C","Xanthine","Xanthosine","Xanthurenic acid","XMP")


Super_class_vec <- c("Organic acids","Carbohydrates","Alkaloids","Organic acids","Nucleic acids","Nucleic acids","Benzenoids","Carbohydrates","Fatty Acyls","Fatty Acyls","Fatty Acyls","Alkaloids","Fatty Acyls","Organic acids","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Fatty Acyls","Fatty Acyls","Organic nitrogen compounds","Fatty Acyls","Fatty Acyls","Benzenoids","Fatty Acyls","Fatty Acyls","Fatty Acyls","Organic nitrogen compounds","Organic acids","Organic acids","Organoheterocyclic compounds","Organic acids","Organic acids","Fatty Acyls","Alkaloids","Nucleic acids","Carbohydrates","Nucleic acids","Fatty Acyls","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organoheterocyclic compounds","Organic acids","Organic acids","Nucleic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Nucleic acids","Organic acids","Organic acids","Organoheterocyclic compounds","Organoheterocyclic compounds","Organoheterocyclic compounds","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Fatty Acyls","Organic acids","Organic nitrogen compounds","Organic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organic nitrogen compounds","Organic nitrogen compounds","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organic acids","Organoheterocyclic compounds","Nucleic acids","Nucleic acids","Nucleic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Carbohydrates","Nucleic acids","Nucleic acids","Organic nitrogen compounds","Organoheterocyclic compounds","Organoheterocyclic compounds","Nucleic acids","Organic acids","Organic acids","Organic acids","Alkaloids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organic acids","Organic acids","Carbohydrates","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organoheterocyclic compounds","Organoheterocyclic compounds","Organic acids","Carbohydrates") 
Super_class_vec <- c(Super_class_vec,"Organic acids","Carbohydrates","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Nucleic acids","Carbohydrates","Carbohydrates","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Carbohydrates","Organic acids","Organic acids","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Organic acids","Nucleic acids","Nucleic acids","Organoheterocyclic compounds","Carbohydrates","Alkaloids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Nucleic acids","Alkaloids","Nucleic acids","Alkaloids","Alkaloids","Nucleic acids","Organic oxygen compounds","Organic acids","Fatty Acyls","Fatty Acyls","Organic acids","Alkaloids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Fatty Acyls","Organic acids","Alkaloids","Organic acids","Organic acids","Organic acids","Organic nitrogen compounds","Alkaloids","Fatty Acyls","Nucleic acids","Organic acids","Organic acids","Alkaloids","Alkaloids","Alkaloids","Organic acids","Organic acids","Alkaloids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Carbohydrates","Organic acids","Alkaloids","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organic acids","Organic acids","Organic acids","Alkaloids","Alkaloids","Alkaloids","Alkaloids","Alkaloids","Organic acids","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Organic acids","Alkaloids","Organic acids","Organic nitrogen compounds","Organic acids","Organic acids","Organoheterocyclic compounds","Nucleic acids","Organoheterocyclic compounds","Nucleic acids","Alkaloids","Alkaloids","Organic acids","Organic acids","Benzenoids","Alkaloids","Organoheterocyclic compounds","Carbohydrates","Carbohydrates","Nucleic acids","Organic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organic acids","Carbohydrates","Organic acids","Organic acids","Alkaloids","Organic acids","Carbohydrates","Alkaloids","Alkaloids","Organic acids","Organic acids","Fatty Acyls","Organic acids","Organic acids","Nucleic acids","Carbohydrates","Organic acids","Nucleic acids","Nucleic acids","Organic acids","Alkaloids","Organic nitrogen compounds","Alkaloids","Organic acids","Organic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Nucleic acids","Organic nitrogen compounds","Organic nitrogen compounds","Nucleic acids","Nucleic acids","Alkaloids","Nucleic acids","Organic acids","Organic acids","Organic acids","Organoheterocyclic compounds","Nucleic acids","Nucleic acids","Alkaloids","Nucleic acids"  )


Main_class_vec <- c("Carboxylic acids","Monosaccharides","Pyridine alkaloids","Amino acids and peptides","Purines","Pyrimidines","Benzenes","Monosaccharides","Fatty acids","Fatty acids","Fatty acids","Pyridine alkaloids","Fatty acids","TCA acids","Short-chain acids","Phenylpropanoids","Purines","Purines","Fatty esters","Fatty esters","Guanidines","Fatty acids","Fatty acids","Benzamides","Fatty acids","Fatty acids","Fatty acids","Amines","Amino acids and peptides","Amino acids and peptides","Pterins","Phenylpropanoids","Phenylpropanoids","Fatty acids","Tryptophan alkaloids","Pyrimidines","Monosaccharides","Purines","Fatty esters","Purines","Purines","Purines","Purines","Purines","Purines","Imidazoles","Imidazoles","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Cyclic ureas","Amino acids and peptides","Amino acids and peptides","Purines","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Purines","Amino acids and peptides","Amino acids and peptides","Bilirubins","Bilirubins","Heterocyclic compounds","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Fatty esters","Amino acids and peptides","Carnitines","Amino acids and peptides","Pyrimidines","Pyrimidines","Pyrimidines","Cholines","Cholines","TCA acids","Amino acids and peptides","Pyrimidines","Pyrimidines","Purines","Amino acids and peptides","Azoles","Pyrimidines","Purines","Pyrimidines","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Pyrimidines","Pyrimidines","Purines","Pyrimidines","Purines","Pyrimidines","Purines","Purines","Monosaccharides","Pyrimidines","Purines","Amines","Pterins","Cyclic ureas","Pyrimidines","Short-chain acids","Amino acids and peptides","Amino acids and peptides","Tyrosine alkaloids","Pyrimidines","Pyrimidines","Pyrimidines","Pyrimidines","Pyrimidines","Pyrimidines","Pyrimidines","Amino acids and peptides","Amino acids and peptides","Monosaccharides","Flavins","Imidazoles","Glycinamidine ribonucleotides","Glycinamide ribonucleotides","Alloxazines and isoalloxazines","Pterins","Amino acids and peptides","Monosaccharides","TCA acids","Monosaccharides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Glycinamide ribonucleotides","Purines","Purines","Monosaccharides","Monosaccharides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Short-chain acids","Monosaccharides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Purines","Purines","Amino acids and peptides","Purines","Purines","Metallotetrapyrroles","Monosaccharides")
Main_class_vec <- c(Main_class_vec,"Histidine alkaloids","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Sulfinic acids","Purines","Histidine alkaloids","Purines","Tryptophan alkaloids","Tryptophan alkaloids","Purines","Alcohols and polyols","Amino acids and peptides","Fatty acids","Fatty acids","Keto acids","Anthranilic acid alkaloids","Amino acids and peptides","Short-chain acids","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Fatty acids","TCA acids","Tryptophan alkaloids","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Guanidines","Histidine alkaloids","Fatty acids","Purines","Amino acids and peptides","Amino acids and peptides","Ornithine alkaloids","Ornithine alkaloids","Ornithine alkaloids","Amino acids and peptides","Amino acids and peptides","Ornithine alkaloids","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Monosaccharides","Amino acids and peptides","Tryptophan alkaloids","Amino acids and peptides","Amino acids and peptides","Nicotinamides","Nicotinamides","Nicotinamides","Nicotinamides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Pyridine alkaloids","Pyridine alkaloids","Pyridine alkaloids","Tryptophan alkaloids","Tryptophan alkaloids","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Pyrimidines","Pyrimidines","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Short-chain acids","Amino acids and peptides","Pyridine alkaloids","Amino acids and peptides","Amines","Amino acids and peptides","Amino acids and peptides","Porphyrins","Pyrimidines","Pterins","Purines","Pyridine alkaloids","Pyridine alkaloids","Amino acids and peptides","Short-chain acids","Benzenes","Pyridine alkaloids","Alloxazines and isoalloxazines","Monosaccharides","Monosaccharides","Pyrimidines","Amino acids and peptides","Purines","Purines","Imidazoles","Amino acids and peptides","Monosaccharides","Amino acids and peptides","Amino acids and peptides","Tryptophan alkaloids","Organic phosphoric acids","Monosaccharides","Ornithine alkaloids","Ornithine alkaloids","Amino acids and peptides","TCA acids","Fatty esters","Amino acids and peptides","Sulfonic acids","Pyrimidines","Monosaccharides","Amino acids and peptides","Pyrimidines","Pyrimidines","Carboxylic acids","Pyridine alkaloids","Amine oxides","Tryptophan alkaloids","Amino acids and peptides","Amino acids and peptides","Pyrimidines","Pyrimidines","Pyrimidines","Pyrimidines","Pyrimidines","Pyrimidines","Organonitrogen compounds","Organonitrogen compounds","Purines","Pyrimidines","Histidine alkaloids","Pyrimidines","Amino acids and peptides","Amino acids and peptides","Amino acids and peptides","Lactones","Purines","Purines","Anthranilic acid alkaloids","Purines" )

Sub_class_vec <- c("Tricarboxylic acids","Hexoses","Nicotinic acid alkaloids","Amino acids","Xanthines","Pyrimidine rNMP","Aminobenzoic acids","Pentoses","Dicarboxylic acids","Hydroxy FA","Dicarboxylic acids","Nicotinic acid alkaloids","Dicarboxylic acids","TCA acids","Short-chain acids","Cinnamic acids","Purine rNMP","Purine rNMP","Acyl carnitines","Acyl CoAs","Guanidines","Dicarboxylic acids","Hydroxy FA","Hippuric acids","Hydroxy FA","Dicarboxylic acids","Oxo FA","Aralkylamines","Amino acids","Amino acids","Biopterins","Cinnamic acids","Cinnamic acids","Amino FA","Indoleacetic acids","Pyrimidones","Amino sugars","Hypoxanthines","Acyl CoAs","Aminopurines","Purine ribonucleosides","Purine rNMP","Purine rNDP","Purine rNDP","Purine rNDP","Imidazole ribonucleosides and ribonucleotides","Imidazole ribonucleosides and ribonucleotides","Dipeptides","Dipeptides","Amino acids","Dipeptides","Dipeptides","Ureides","Amino acids","Amino acids","Purine rNMP","Dipeptides","Amino acids","Amino acids","Dipeptides","Amino acids","Amino acids","Purine rNTP","Amino acids","Amino acids","Bilirubins","Bilirubins","Biotins","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Acyl carnitines","Amino acids","Carnitines","Dipeptides","Pyrimidine rNDP","Pyrimidine rNDP","Pyrimidine rNDP","Cholines","Cholines","TCA acids","Amino acids","Pyrimidine rNMP","Pyrimidine ribonucleosides","Other purines","Amino acids","Imidazolines","Pyrimidine rNTP","Purine ribonucleosides","Pyrimidine rNMP","Dipeptides","Amino acids","Amino acids","Amino acids","Dipeptides","Amino acids","Pyrimidine ribonucleosides","Pyrimidones","Purine dNMP","Pyrimidine dNMP","Purine deoxyribonucleosides","Pyrimidine deoxyribonucleosides","Purine deoxyribonucleosides","Purine deoxyribonucleosides","Pentose phosphates","Pyrimidine deoxyribonucleosides","Purine dNMP","1,2-aminoalcohols","Folic acids","Ureides","Other pyrimidines","Short-chain acids","Amino acids","Amino acids","Phenylethylamines","Pyrimidine dNDP","Pyrimidine ribonucleosides","Pyrimidine dNMP","Pyrimidine dNTP","Pyrimidine dNMP","Pyrimidine dNTP","Hydropyrimidines","Dipeptides","Amino acids","Sugar acids","Flavin nucleotides","Imidazole ribonucleosides and ribonucleotides","Glycinamidine ribonucleotides","Glycinamide ribonucleotides","Flavins","Folic acids","Amino acids","Hexose phosphates","TCA acids","Sugar acids","Dipeptides","Dipeptides","Dipeptides","Dipeptides","Dipeptides") 
Sub_class_vec <- c(Sub_class_vec,"Glycinamide ribonucleotides","Purine rNDP","Purine rNDP","Sugar acids","Sugar acids","Amino acids","Amino acids","Amino acids","Tripeptides","Tripeptides","Dipeptides","Short-chain acids","Sugar acids","Amino acids","Dipeptides","Dipeptides","Dipeptides","Purine rNMP","Purine rNTP","Amino acids","Hypoxanthines","Purine ribonucleosides","Metalloporphyrins","Hexose phosphates","Imidazole alkaloids","Amino acids","Dipeptides","Amino acids","Amino acids","Amino acids","Amino acids","Amino acids","Amino acids","Sulfinic acids","Hypoxanthines","Imidazole alkaloids","Purine rNMP","Indoleacetic acids","Simple indole alkaloids","Purine ribonucleosides","Inositols","Amino acids","Dicarboxylic acids","Branched FA","Short-chain keto acids","Acridone alkaloids","Amino acids","Short-chain acids","Amino acids","Amino acids","Dipeptides","Dicarboxylic acids","TCA acids","Tryptamines","Amino acids","Amino acids","Amino acids","Guanidines","Imidazole alkaloids","Dicarboxylic acids","Purine ribonucleosides","Amino acids","Amino acids","Polyamines","Polyamines","Polyamines","Amino acids","Amino acids","Polyamines","Amino acids","Amino acids","Amino acids","Dipeptides","Amino acids","Amino acids","Amino acids","Sugar acids","Amino acids","Tryptamines","Amino acids","Amino acids","Nicotinamide dinucleotides","Nicotinamide dinucleotides","Nicotinamide dinucleotides","Nicotinamide dinucleotides","Amino acids","Amino acids","Amino acids","Nicotinic acid alkaloids","Nicotinic acid alkaloids","Nicotinic acid alkaloids","Tryptamines","Tryptamines","Amino acids","Tripeptides","Amino acids","Pyrimidine carboxylic acids","Pyrimidine ribonucleosides","Amino acids","Amino acids","Amino acids","Amino acids","Amino acids","Short-chain acids","Amino acids","Nicotinic acid alkaloids","Amino acids","Aralkylamines","Amino acids","Amino acids","Porphyrins","Pyrimidine ribonucleosides","Pterins","Other purines","Nicotinic acid alkaloids","Nicotinic acid alkaloids","Amino acids","Short-chain acids","Hydroxybenzoic acids","Nicotinic acid alkaloids","Flavins","Sugar acids","Pentose phosphates","Pyrimidine ribonucleosides","Amino acids","Other purines","Other purines","Imidazole ribonucleosides and ribonucleotides","Amino acids","Monosaccharide phosphates","Dipeptides","Amino acids","Tryptamines","Organic phosphoric acids","Sugar alcohols","Polyamines","Polyamines","Amino acids","TCA acids","Acyl CoAs","Amino acids","Sulfonic acids","Thiamines","Sugar acids","Amino acids","Pyrimidine deoxyribonucleosides","Other pyrimidines","Tricarboxylic acids","Nicotinic acid alkaloids","Trisubstituted amine oxides","Tryptamines","Amino acids","Amino acids","Pyrimidine rNDP","Pyrimidine rNDP","Pyrimidine rNDP","Pyrimidine rNDP","Pyrimidine rNMP","Other pyrimidines","Ureas","Ureas","Xanthines","Pyrimidine ribonucleosides","Imidazole alkaloids","Pyrimidine rNTP","Dipeptides","Dipeptides","Amino acids","Gamma butyrolactones","Xanthines","Purine ribonucleosides","Acridone alkaloids","Purine rNMP" )

HMDB_ID_vec <- c("HMDB0031193","HMDB0002712","HMDB0000699","HMDB0000001","HMDB0010738","HMDB0011691","HMDB0001123","HMDB0003224","HMDB0000321","HMDB0000008","HMDB0059655","HMDB0013751","HMDB0000225","HMDB0000208","HMDB0003391","HMDB0002643","HMDB0000058","HMDB0001314","HMDB0001161","HMDB0001373","HMDB0013222","HMDB0000345","HMDB0000357","HMDB0006116","HMDB0000023","HMDB0000355","HMDB0000005","HMDB0001861","HMDB0000479","HMDB0000528","HMDB0002215","HMDB0000755","HMDB0000707","HMDB0001149","HMDB0000763","HMDB0002894","HMDB0001128","HMDB0000897","HMDB0001206","HMDB0000034","HMDB0000050","HMDB0000536","HMDB0001341","HMDB0006557","HMDB0001178","HMDB0062179","HMDB0001235","HMDB0028685","HMDB0028686","HMDB0000161","HMDB0028697","HMDB0028699","HMDB0000462","HMDB0301759","NA","HMDB0000045","HMDB0000194","HMDB0000517","HMDB0000052","HMDB0028736","HMDB0000168","HMDB0000191","HMDB0000538","HMDB0000056","HMDB0000043","HMDB0000054","HMDB0001008","HMDB0000030","HMDB0240771","HMDB0062514","HMDB0000736","HMDB0000688","HMDB0013127","HMDB0002366","HMDB0013128","HMDB0002095","HMDB0061717","HMDB0013130","HMDB0061677","HMDB0000828","HMDB0000062","HMDB0000033","HMDB0001546","HMDB0001413","HMDB0001564","HMDB0000097","HMDB0001565","HMDB0000094","HMDB0000904","HMDB0000095","HMDB0001176","HMDB0001423","HMDB0000064","HMDB0000562","HMDB0000082","HMDB0249529","NA","HMDB0000078","HMDB0000099","HMDB0303991","HMDB0000731","HMDB0000709","HMDB0000192","HMDB0000089","HMDB0000630","HMDB0000905","HMDB0001202","HMDB0000101","HMDB0000014","HMDB0000085","HMDB0000071","HMDB0001031","HMDB0000012","HMDB0001044","HMDB0004437","HMDB0001056","HMDB0304210","HMDB0000076","HMDB0001473","HMDB0001539","HMDB0000092","HMDB0000073","HMDB0001274","HMDB0001328","HMDB0001227","HMDB0001342","HMDB0001409","HMDB0001191","HMDB0240650","HMDB0003869","HMDB0003045","HMDB0000613","HMDB0001248","HMDB0001439","HMDB0002305","HMDB0001308","HMDB0001520","HMDB0000121","HMDB0000854","HMDB0003971","HMDB0000134","HMDB0002545","HMDB0001049","HMDB0011737","HMDB0011738","HMDB0029151","HMDB0011172","HMDB0002022","HMDB0001201","HMDB0001095","HMDB0000663","HMDB0000127","HMDB0000148","HMDB0000641","HMDB0000590","HMDB0000125","HMDB0000125","HMDB0028837","HMDB0001112","HMDB0000139","HMDB0000123","HMDB0028844","HMDB0000721","HMDB0028853","HMDB0001397","HMDB0001273","HMDB0000128","HMDB0000132","HMDB0000133","HMDB0003178","HMDB0246205","HMDB0000870","HMDB0000177","HMDB0028898","HMDB0000670","HMDB0000679","HMDB0000742","HMDB0000719","HMDB0011631","HMDB0240251","HMDB0000965","HMDB0000157","HMDB0002024","HMDB0000175","HMDB0000197","HMDB0000682") 
HMDB_ID_vec <- c(HMDB_ID_vec,"HMDB0000195","NA","HMDB0000172","HMDB0002092","HMDB0000019","HMDB0000695","HMDB0000715","HMDB0000684","HMDB0000190","HMDB0000687","HMDB0000182","HMDB0028964","HMDB0000176","HMDB0000156","HMDB0001389","HMDB0000696","HMDB0062174","HMDB0002005","HMDB0001522","HMDB0002820","HMDB0000202","NA","HMDB0000446","HMDB0002038","HMDB0002172","HMDB0041947","HMDB0001186","HMDB0000206","HMDB0001325","HMDB0002189","HMDB0000766","HMDB0006028","HMDB0000812","HMDB0001067","HMDB0001138","HMDB0006029","HMDB0032055","HMDB0000230","HMDB0003357","HMDB0001238","HMDB0062557","HMDB0013713","HMDB0000902","HMDB0001487","HMDB0304435","HMDB0000221","HMDB0000828","HMDB0255145","HMDB0240317","HMDB0001406","HMDB0000229","HMDB0001488","HMDB0004369","HMDB0004370","HMDB0003011","HMDB0005765","HMDB0000214","HMDB0000226","HMDB0000788","HMDB0255868","NA","HMDB0000210","HMDB0000159","HMDB0001511","HMDB0000263","HMDB0000272","HMDB0002243","HMDB0000716","HMDB0000245","HMDB0000162","HMDB0004827","HMDB0001097","HMDB0000767","HMDB0000802","HMDB0001366","HMDB0001545","HMDB0000239","HMDB0000267","HMDB0000243","HMDB0003072","HMDB0000232","HMDB0000244","HMDB0000867","HMDB0001548","HMDB0000884","HMDB0000279","HMDB0000939","HMDB0001185","HMDB0000797","HMDB0000271","HMDB0258206","HMDB0029032","HMDB0000187","HMDB0000259","HMDB0000126","HMDB0000247","HMDB0001257","HMDB0001256","NA","HMDB0000254","HMDB0001022","HMDB0029416","HMDB0000251","HMDB0000235","HMDB0000943","HMDB0000167","HMDB0000273","HMDB0000262","HMDB0000958","HMDB0000875","HMDB0000925","HMDB0000303","HMDB0000929","HMDB0000158","HMDB0000295","HMDB0000286","HMDB0000935","HMDB0000290","HMDB0000288","HMDB0000300","HMDB0000294","HMDB0000026","HMDB0000289","HMDB0000296","HMDB0000301","HMDB0000285","HMDB0029123","HMDB0029126","HMDB0000883","HMDB0000044","HMDB0000292","HMDB0000299","HMDB0000881","HMDB0001554" )

KEGG_ID_vec <- c("C19806","C07326","C02918","C01152","C16358","C02354","C00108","C01801","C02360","C05984","C02630","C02502","C00322","C00026","C00631","NA","C00575","C00942","C01181","C00882","C03065","NA","C01089","NA","C01188","C03761","C00109","NA","C01152","NA","C00268","C03672","C01179","C00430","C05635","C02376","C03090","C02242","C00024","C00147","C00212","C03794","C00008","C00498","C00301","NA","C03373","NA","NA","C00041","NA","NA","NA","C08265","NA","C00020","C01262","C00062","C03406","NA","C00152","C00049","C00002","C00099","NA","C00486","C00500","C00120","NA","C03017","NA","NA","NA","NA","NA","NA","NA","NA","NA","C00438","C00487","C00386","C00112","C00307","C00570","C00114","NA","C00158","C00327","C00055","C00128","C00010","C00300","C00791","C00063","NA","NA","C01419","C02291","C00506","C05824","NA","C00491","C00475","C00380","C00360","C00239","C00559","C00881","C00330","C05512","C00673","C00526","C00362","C06772","C00415","C00906","C00429","C00111","C03626","C01026","C03758","C00363","C00842","C00364","C00459","C00365","C00460","C06231","NA","C05570","NA","C00016","C04734","C04640","C04376","C00061","C00504","C00439","C00085","C00122","C08348","C00669","C05282","C05283","NA","NA","C03838","C00035","C00325","C00818","C00191","C00025","C00064","NA","C00051","C00051","NA","C00118","C00258","C00037","NA","NA","NA","C00144","C00044","C00581","C00242","C00387","C00032","C02965","C00388","C00135","NA","C01924","C02427","C00155","C00263","C03227","C01015","C00519","C00262","C02835","C00130","C00954","NA","C00294","NA","C00407","C00490","C00141","C00233","C01717","C00328","C00186","C00123","C00047","NA","C01384","C00149","C01598","C00073","NA","NA","C02294","C05828","C02170","NA","C12989","C02728","C03413","NA","C02567","C02727","C03793","C01029","NA","NA","C01042","C12270","C00624","NA","C02997","C19910","C00437","C00978","NA","NA","C00003","C00004","C00006","C00005","NA","NA","NA","C00153","C00455","C00253","C06212","C06213","C00979","NA","C00077","C00295","C01103","C01118","NA","NA","C00079","C02305","C00074","C01005","C10164","C00408","C00931","C00148","C10172","C01079","C02067","C00715","C15587","C00250","C00314","C01879","C00022","C00296","C03722","C00255","C01685","C00117","NA","C00449","C00021","C00019","C04823","C00213","C05382","NA","C00065","C00780","C00093","C00794","C00315","C00750","NA","C00042","C00091","C03884","C00245","C00378","C01620","C00188","C00214","C00178","C02341","C01004","C01104","C00398","C00078","C00082","C00015","C00029","C00167","C00043","C00105","C00106","C00086","C02642","C00366","C00299","C00785","C00075","NA","NA","C00183","C00072","C00385","NA","C02470","C00655") 


universe <- data.frame(
  'Input name' = Input_name_vec,
  'Super class' = Super_class_vec,
  'Main class' = Main_class_vec,
  'Sub class' = Sub_class_vec,
  'HMDB_ID' = HMDB_ID_vec,
  'KEGG_ID' = KEGG_ID_vec,
  stringsAsFactors = FALSE
)


# to make it first get the refmet txt and loadit in r calles universe then 
# a <- unlist(universe$KEGG_ID) |> as.vector()
# writeLines(paste0('"', a, '"', collapse = ","), "input_names.csv")
# open the file with notepad++, copy paste in r

print("#0.3. Set universe finished!")
```

#0.4. Load Data.
```{r}
#1. Load Sample Data
file_listTF <- list.files(input.dir, pattern = ".xlsx", full.names = TRUE, recursive = TRUE) # make a list of xls files in TF folder folder
if(sum(grep("~", file_listTF))>0){file_listTF <- file_listTF[-grep("~", file_listTF)]} # remove any open xls
suppressMessages(data <- readxl::read_excel(file_listTF[grep("ssf", tolower(file_listTF), invert = T)], sheet = 1, col_names = T) |> as.data.frame()) # load data file
rownames(data) <- data[[1]]   # set first column as row names
data <- data[, -1]    # Remove first col

data[] <- lapply(data, function(col) {as.numeric(gsub(" ", "", col)) })# remove spaces and make numeric
if(unique(sapply(data, is.numeric))==FALSE){warning("Input data are NOT numeric!!")} # Check if data are indeed numeric

data[is.na(data)] <- 0 # Set all NAs to zeros


########  ### ### CHANGE DATA ROWNAMES HERE ------------

# Run each row above up to here and then make changes to the raw names so that the ones in the SSF match exactly the rownames of the data.

#stop

#2. Load Sample SSF
suppressMessages(SSF <- readxl::read_excel(file_listTF[grep("ssf", tolower(file_listTF))], sheet = 1, col_names = T) %>% as.data.frame()) # load SSF file
sel.col <- grep("Vial Identifier", SSF, ignore.case = FALSE, value = FALSE)  # find the column we want
sel.row <- grep("Vial Identifier", t(SSF[,grep("Vial Identifier", SSF, ignore.case = FALSE, value = FALSE)]), ignore.case = FALSE, value = FALSE)  # find the row we want
SSF <- SSF[sel.row:length(as.data.frame(t(SSF))),sel.col:ncol(SSF)] # subset the data
if (sum(is.na(SSF[,1])) != 0){  SSF <- SSF[1:min(which(is.na(SSF[,1]))) - 1,]  } # Remove NA rows in the end
colnames(SSF) <- SSF[1,] # make first row into col names
SSF <- SSF[-c(1:4),] # remove the first 4 rows with information
if(sum(grep("Example", SSF[1,]))>0 ){
  SSF <- SSF[-1,] # remove the first 4 rows with information
}
rownames(SSF) <- SSF[,1]  # make 1st column into row names
SSF <- SSF[,-1] # delete fist column
colnames(SSF)[1] <- "Conditions"

SSF$Conditions <- factor(SSF$Conditions, levels = unique(SSF$Conditions))

########  ### ### CHANGE SSF ROWNAMES HERE ---------------


# Run each row above up to here and then make changes to the raw names so that the ones in the SSF match exactly the rownames of the data.

#stop


# Add extra samples in the SSF
# Remove any pools or blanks from the SSF
SSF <- SSF[!grepl("pool", tolower(rownames(SSF)),ignore.case=TRUE ) ,] 
SSF <- SSF[!grepl(blank_samp_id, tolower(rownames(SSF)),ignore.case=TRUE ) ,] 


# Add pool samples back in the SSF
Pool_metadata <- data.frame(matrix(NA, nrow = sum(grepl("Pool",rownames(data),ignore.case=TRUE)), ncol = ncol(SSF))) # Name a dataframe for pool metadata
colnames(Pool_metadata) <- colnames(SSF) # add colnames from metadata
rownames(Pool_metadata) <-  rownames(data)[grepl("Pool", rownames(data),ignore.case=TRUE)] # add rownames of the pool samples from the data
Pool_metadata$Conditions <- rownames(Pool_metadata)
Pool_metadata$Conditions <-   "Pool"


# Add Blank samples back in the SSF
Blank_metadata <- data.frame(matrix(NA, nrow = sum(grepl(blank_samp_id,rownames(data),ignore.case=TRUE)), ncol = ncol(SSF))) # Name a dataframe for pool metadata
colnames(Blank_metadata) <- colnames(SSF) # add colnames from metadata
rownames(Blank_metadata) <-  rownames(data)[grepl(blank_samp_id, rownames(data),ignore.case=TRUE)] # add rownames of the pool samples from the data
Blank_metadata$Conditions <- rownames(Blank_metadata)
Blank_metadata$Conditions <- "Blank"


# Add in a column multiple thingis
# Pool_metadata$`Sample Matrix` <- dplyr::case_when(
#   grepl("Heart", rownames(Pool_metadata), ignore.case = TRUE) ~ "HEART",
#   grepl("Kidney", rownames(Pool_metadata), ignore.case = TRUE) ~ "Kidney",
#   grepl("SI", rownames(Pool_metadata), ignore.case = TRUE) ~ "small intestine",
#   grepl("Muscle", rownames(Pool_metadata), ignore.case = TRUE) ~ "MUSCLE",
#   TRUE ~ "UNKNOWN" # Default value if no match
# )

#  merge the metadata with the pool metadata
SSF <- rbind(SSF, Pool_metadata, Blank_metadata) 
rm(Pool_metadata,Blank_metadata)

# Subset SSF to keep only sample names that exist in the data, as actual sample number might be different from the originally planned.
SSF <- SSF[rownames(data), , drop = FALSE]

# Extract growth vector
if(CoRe==TRUE){
  ssf_for_growth_factor <- SSF %>% select(Conditions, `Cell Number at Time Zero`  ,`Cell Count at Experimental End Point` )
  ssf_for_growth_factor$`Cell Number at Time Zero` <- 1
  ssf_for_growth_factor <- ssf_for_growth_factor[complete.cases(ssf_for_growth_factor),]

  ssf_for_growth_factor[, 2:3] <- lapply(ssf_for_growth_factor[, 2:3], as.numeric)
  growth_factor <- ssf_for_growth_factor %>%
  mutate(GrowthFactor = `Cell Count at Experimental End Point` /
                   `Cell Number at Time Zero` )%>%
  group_by(Conditions) %>%
  summarise(GrowthFactor = unique(GrowthFactor), .groups = "drop")
  
  growth_factor
}

# Keep only columns that are NOT all NA
SSF <- SSF[, colSums(!is.na(SSF)) > 0]


#3. Sample Run Order
file_listRunOrder <- list.files(input.dir, pattern = ".csv",recursive = TRUE, full.names = TRUE) # get all csv files in the raw data folder and subfolders
if(sum(grep("~", file_listRunOrder))>0){file_listRunOrder <- file_listRunOrder[-grep("~", file_listRunOrder)]} # remove open xls

RandomizedSamples <- read.csv(file_listRunOrder) %>% as.data.frame() 
colnames(RandomizedSamples) <- RandomizedSamples[1,] # make first row into col names
RandomizedSamples <- RandomizedSamples[-c(1),] # remove the first 4 rows with information
RandomizedSamples <- RandomizedSamples[,1]



### ### ### Here make changes to the Run Order --------------

# RandomizedSamples[grep(blank_samp_id, RandomizedSamples)][1:length( rownames(data)[grep(blank_samp_id, rownames(data))])] <- rownames(data)[grep(blank_samp_id, rownames(data))]
#stop


RandomizedSamples <- RandomizedSamples[RandomizedSamples %in% rownames(SSF)]

SSF$RunOrder <- match(rownames(SSF), RandomizedSamples)
SSF$RunOrder  <- SSF$RunOrder - min(SSF$RunOrder)+1
rm(RandomizedSamples)


# Remove first 5 pools( Pools that are used to condition the LC column) as they can be incosistent
if(length(grep("PoolB", rownames(data)))>1){
  starting_pools_to_remove <- c("PoolA00", "PoolA01" , "PoolA02")#, "PoolA03" )
}
if(length(grep("PoolB", rownames(data)))>1){
  starting_pools_to_remove <- c("PoolB00", "PoolB01" , "PoolB02")#, "PoolB03" )
}
if(length(grep("PoolC", rownames(data)))>1){
  starting_pools_to_remove <- c("PoolC00", "PoolC01" , "PoolC02")#, "PoolC03" )
}


rownames(data)

data <- data[-grep(paste(starting_pools_to_remove, collapse="|"), rownames(data)),]

# Subset SSF to keep only sample names that exist in the data, as actual sample number might be different from the originally planned.
SSF <- SSF[rownames(data), , drop = FALSE]
# Keep only columns that are NOT all NA
SSF <- SSF[, colSums(!is.na(SSF)) > 0]


SSF_raw <- SSF
data_raw <- data
# Input checks, if all data have the same samples. 
message("All samples in SSF found in data: ", all(rownames(SSF) %in% rownames(data)))
message("All samples in data found in SSF: ", all(rownames(data) %in% rownames(SSF)))

### ----- Check the data ----- ###
# Check the columns that have NAs
check <- data[-grep(blank_samp_id, tolower(rownames(data))),]
Check <- check %>% select( names(check)[colSums(check==0) > 0])

message("There are ",length(colnames(Check)) , " metabolites with missing values." )
# check sample groups
message("Number of pool samples is:", length(grep("pool", tolower(SSF$Conditions))))
message("Number of the smallest sample group is: ",SSF %>% count(Conditions) %>% arrange(n) %>% slice(1)%>% select(n))

print("#0.4. Load Data finished!")
```


#0.5. Set colors.
```{r}

### ---------- Create Custom Colors for plotting ---------- ###
# Check sample groups to select colors.
length(unique(SSF$Conditions) )
unique(SSF$Conditions)

# Create customColors table
customColors <- data.frame(
  Conditions = unique(SSF$Conditions),
  # colors <- c("red", "blue", "green", "orange", "purple", "cyan", "magenta","white, "grey")[1:length(unique(SSF$Conditions))]
  Color = c( "#8eb841", "red", "blue", "grey","black", "#3c4f1b", "red","black","lightblue", "blue", "darkblue", "#bdd78d", "#8eb841", "#3c4f1b", "red" ,"grey")[1:length(unique(SSF$Conditions))]
)
customColorsBox <- customColors

# Map colors to SSF without merging
SSF$Color <- customColors$Color[match(SSF$Conditions, customColors$Conditions)]
SSF_raw <- SSF

# Set the Media_sample and their controls.
if(CoRe==T){
  media_sample_names <- rownames(SSF)[-grep("Ctrl", rownames(SSF))]
  media_control_names <- c(rownames(SSF)[grep("Ctrl", rownames(SSF))],rownames(SSF)[grep("Blank", rownames(SSF))])
}
print("#0.5. Set colors!")
```

#0.6. Load helper functions.
```{r}
# res <- make_pca_grid(data, SSF, out_dir = Data_cleaning_dir)
make_pca_grid <- function(data, SSF,
                          add_to_name = NULL,
                          project_code = ProjectCode,
                          color_col = "Conditions",
                          fig_size_cm = c(60, 40)) {
  
  suppressMessages(suppressWarnings({
    
    PCA_dir <- file.path(output.dir, "PCA_plots")
    if (!dir.exists(PCA_dir)) dir.create(PCA_dir, recursive = TRUE)
    
    # labels
    if (!is.null(project_code)) {
      SSF$Label <- sub(paste0("^", project_code, "_"), "", rownames(SSF))
    } else {
      SSF$Label <- rownames(SSF)
    }
    
    # PCA input: copy, fill NAs with 0, drop zero-variance features
    X <- data
    X[is.na(X)] <- 0
    nzv <- apply(X, 2, sd) > 0
    if (!all(nzv)) X <- X[, nzv, drop = FALSE]
    if (ncol(X) < 2) stop("Not enough non-zero-variance features for PCA.")
    
    # PCA
    pca_res <- prcomp(X, scale. = TRUE)
    n_pc_use <- min(5, ncol(pca_res$x))
    
    # palette from SSF if available
    pal <- NULL
    if ("Color" %in% colnames(SSF) && color_col %in% colnames(SSF)) {
      pal_df <- unique(SSF[, c(color_col, "Color")])
      pal <- setNames(pal_df$Color, pal_df[[color_col]])
    }
    
    # builder
    build_grid <- function(label_points = FALSE) {
      plots <- vector("list", n_pc_use * n_pc_use)
      k <- 1
      for (i in seq_len(n_pc_use)) {
        for (j in seq_len(n_pc_use)) {
          if (i != j) {
            p <- autoplot(
              pca_res, data = SSF, x = i, y = j,
              colour = color_col, size = 2, alpha = 0.8, loadings = FALSE
            ) +
              theme_classic() +
              ggtitle(paste("PC", i, "vs PC", j)) +
              geom_hline(yintercept = 0, color = "black", linewidth = 0.1) +
              geom_vline(xintercept = 0, color = "black", linewidth = 0.1)
            if (!is.null(pal)) p <- p + scale_color_manual(values = pal)
            if (label_points) p <- p + ggrepel::geom_text_repel(aes(label = Label), size = 2.5)
          } else {
            p <- ggplot() + theme_void() + ggtitle(paste0("PC", i, " vs ", j))
          }
          plots[[k]] <- p; k <- k + 1
        }
      }
      patchwork::wrap_plots(plots, ncol = n_pc_use)
    }
    
    # build + save
    p_plain   <- build_grid(FALSE)
    p_labeled <- build_grid(TRUE)
    
    ggsave(file.path(PCA_dir, paste0(add_to_name, "__PCA.svg")),
           plot = p_plain, width = fig_size_cm[1], height = fig_size_cm[2], units = "cm")
    ggsave(file.path(PCA_dir,  paste0(add_to_name, "__PCA_Labeled.svg")),
           plot = p_labeled, width = fig_size_cm[1], height = fig_size_cm[2], units = "cm")
    
    invisible(list(plain = p_plain, labeled = p_labeled, pca = pca_res))
  }))
}

make_signal_drift_reports <- function(data, SSF,
                                      add_to_name = NULL,
                                      project_code = ProjectCode,
                                      valine_pattern = valine_pattern_id) {
  
  # silence warnings/messages inside this function
  old_warn <- getOption("warn"); options(warn = -1)
  on.exit(options(warn = old_warn), add = TRUE)
  
  suppressMessages(suppressWarnings({
    
    SignalDrift_dir <- file.path(output.dir, "SignalDrift")
    if (!dir.exists(SignalDrift_dir)) dir.create(SignalDrift_dir, recursive = TRUE)
    
    # Palette from SSF
    pal_df <- unique(SSF[, c("Conditions","Color")])
    customColors <- setNames(pal_df$Color, pal_df$Conditions)
    
    # Pool label (character)
    v_cond <- as.character(SSF$Conditions)
    pool_label <- {
      u <- unique(v_cond[grepl("pool", tolower(v_cond))])
      if (length(u)) u[1] else NA_character_
    }
    
    # Order by RunOrder
    valid_samples <- rownames(SSF)[order(SSF$RunOrder)]
    
    # ---- Prepare Drift Data ----
    Drift_data <- merge(SSF %>% dplyr::select(Conditions), data, by = 0, sort = FALSE)
    colnames(Drift_data)[1] <- "File.Name"
    rownames(Drift_data) <- Drift_data$File.Name
    
    Drift_data_ordered <- Drift_data[match(valid_samples, Drift_data$File.Name), , drop = FALSE]
    Drift_data_ordered <- Drift_data_ordered[!is.na(rownames(Drift_data_ordered)), ]
    Drift_data_ordered$File.Name <- NULL
    Drift_data_ordered[, -1] <- sapply(Drift_data_ordered[, -1], as.numeric)
    
    row_medians <- matrixStats::rowMedians(as.matrix(Drift_data_ordered[, -1]), na.rm = TRUE)
    row_sums    <- rowSums(as.matrix(Drift_data_ordered[, -1]), na.rm = TRUE)
    
    plotdata <- data.frame(SampleMedian = row_medians,
                           TIC = row_sums,
                           SampleGroup = Drift_data_ordered$Conditions) %>%
      tibble::rownames_to_column("RowNames")
    plotdata$RowNames <- factor(plotdata$RowNames, levels = plotdata$RowNames)
    plotdata$Order <- seq_len(nrow(plotdata))
    
    # palette for used groups
    pal_for <- function(groups) {
      used <- intersect(names(customColors), unique(as.character(groups)))
      customColors[used]
    }
    
    # file prefix
    prefix <- if (!is.null(add_to_name) && nzchar(add_to_name)) paste0(add_to_name, "__") else ""
    
    # ---- Plot helper ----
    make_drift_plot <- function(df, yvar, ylab, title, filename) {
      pal_use <- pal_for(df$SampleGroup)
      df$SampleGroup <- factor(as.character(df$SampleGroup), levels = names(pal_use))
      pool_df <- if (!is.na(pool_label)) df[as.character(df$SampleGroup) == pool_label, , drop = FALSE] else NULL
      
      p <- ggplot(df, aes(x = RowNames, y = .data[[yvar]], fill = SampleGroup)) +
        geom_col(position = "dodge") +
        geom_smooth(
          data = pool_df,
          aes(x = Order, y = .data[[yvar]]),
          method = "lm", se = FALSE, color = "red", inherit.aes = FALSE
        ) +
        scale_fill_manual(values = pal_use, drop = FALSE) +
        labs(title = title, x = "Samples by acquirement order", y = ylab) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 40, hjust = 1),
              plot.title = element_text(size = 22))
      
      ggsave(file.path(SignalDrift_dir, paste0(prefix, filename)),
             plot = p, width = 10, height = 8)
    }
    
    # ---- Signal Drift Plots ----
    make_drift_plot(plotdata, "SampleMedian", "Sample Medians",
                    "Signal drift, Sample Median", "SignalDrift_Median_SampleGroup.svg")
    make_drift_plot(plotdata, "TIC", "Sample TIC",
                    "Signal drift, TIC", "SignalDrift_TIC_SampleGroup.svg")
    
    plotdata_pool <- transform(plotdata,
                               SampleGroup = if (!is.na(pool_label)) ifelse(SampleGroup == pool_label, pool_label, "Sample") else "Sample")
    make_drift_plot(plotdata_pool, "SampleMedian", "Sample Medians",
                    "Signal drift, Median, Pool Samples", "SignalDrift_Median_PoolSamples.svg")
    make_drift_plot(plotdata_pool, "TIC", "Sample TIC",
                    "Signal drift, TIC, Pool Samples", "SignalDrift_TIC_PoolSamples.svg")
    
    # ---- Valine-d8 block (optional; runs only if found) ----
    valine_idx <- grep(valine_pattern, tolower(colnames(Drift_data_ordered)), ignore.case = TRUE)
    valine_done <- FALSE
    
    if (length(valine_idx) > 0) {
      valine_col <- colnames(Drift_data_ordered)[valine_idx[1]]
      plotdataIS <- data.frame(
        ValineIS    = Drift_data_ordered[[valine_col]],
        SampleGroup = Drift_data_ordered$Conditions,
        RowNames    = rownames(Drift_data_ordered),
        check.names = FALSE
      )
      plotdataIS$Index <- seq_len(nrow(plotdataIS))
      plotdataIS$RowNames <- factor(plotdataIS$RowNames,
                                    levels = rownames(Drift_data_ordered), ordered = TRUE)
      pal_use <- pal_for(plotdataIS$SampleGroup)
      
      val_cv <- function(x) round(100 * sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE), 2)
      ValSampCV <- val_cv(plotdataIS$ValineIS)
      ValPoolCV <- if (!is.na(pool_label)) val_cv(plotdataIS$ValineIS[plotdataIS$SampleGroup == pool_label]) else NA_real_
      
      p_val <- ggplot(plotdataIS, aes(x = RowNames, y = ValineIS, fill = SampleGroup)) +
        geom_col(position = "dodge") +
        geom_smooth(aes(x = Index, y = ValineIS), method = "lm", se = FALSE, color = "red", inherit.aes = FALSE) +
        scale_y_continuous(limits = c(0, NA)) +
        scale_fill_manual(values = pal_use, drop = FALSE) +
        labs(title = paste0("Signal drift, ", valine_col, " (IS)"),
             subtitle = paste("PoolCV =", ValPoolCV, ", SampleCV =", ValSampCV),
             x = "Samples by acquirement order", y = "Peak Area") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 40, hjust = 1),
              plot.title = element_text(size = 22))
      ggsave(file.path(SignalDrift_dir, paste0(add_to_name, "_SignalDrift_Valine-d8(IS).svg")),
             plot = p_val, width = 12, height = 10)
      
      # Correlations
      plotdata_combined <- merge(plotdata, plotdataIS %>% dplyr::select(-SampleGroup), by = "RowNames")
      cor_TIC_valine    <- cor(plotdata_combined$TIC,          plotdata_combined$ValineIS, use = "complete.obs")
      cor_median_valine <- cor(plotdata_combined$SampleMedian, plotdata_combined$ValineIS, use = "complete.obs")
      
      # Extraction Buffer (optional)
      eb_col <- grep("^Extraction\\s*Buffer$", names(SSF), ignore.case = TRUE, value = TRUE)
      if (length(eb_col) == 1) {
        plotdata_combined$ExtractionBuffer <- SSF[match(plotdata_combined$RowNames, rownames(SSF)), eb_col]
        plotdata_combined$ExtractionBuffer <- as.numeric(plotdata_combined$ExtractionBuffer)
        plotdata_combined$ExtractionBuffer[is.na(plotdata_combined$ExtractionBuffer)] <-
          mean(plotdata_combined$ExtractionBuffer, na.rm = TRUE)
        cor_ExtrBuffer_valine <- cor(plotdata_combined$ExtractionBuffer, plotdata_combined$ValineIS, use = "complete.obs")
      } else {
        plotdata_combined$ExtractionBuffer <- NA_real_
        cor_ExtrBuffer_valine <- NA_real_
      }
      
      p1 <- ggplot(plotdata_combined, aes(x = ValineIS, y = TIC, color = SampleGroup)) +
        geom_point(size = 3) + geom_smooth(method = "lm", se = FALSE, color = "black") +
        scale_color_manual(values = pal_for(plotdata_combined$SampleGroup), drop = FALSE) +
        labs(title = paste("TIC vs", valine_col, "(r =", round(cor_TIC_valine, 3), ")"),
             x = paste(valine_col, "(IS Peak Area)"), y = "TIC") + theme_minimal()
      
      p2 <- ggplot(plotdata_combined, aes(x = ValineIS, y = SampleMedian, color = SampleGroup)) +
        geom_point(size = 3) + geom_smooth(method = "lm", se = FALSE, color = "black") +
        scale_color_manual(values = pal_for(plotdata_combined$SampleGroup), drop = FALSE) +
        labs(title = paste("Sample Median vs", valine_col, "(r =", round(cor_median_valine, 3), ")"),
             x = paste(valine_col, "(IS Peak Area)"), y = "Sample Median") + theme_minimal()
      
      p3 <- ggplot(plotdata_combined, aes(x = ValineIS, y = ExtractionBuffer, color = SampleGroup)) +
        geom_point(size = 3) + geom_smooth(method = "lm", se = FALSE, color = "black") +
        scale_color_manual(values = pal_for(plotdata_combined$SampleGroup), drop = FALSE) +
        labs(title = paste("Extraction Buffer vs", valine_col, "(r =", round(cor_ExtrBuffer_valine, 3), ")"),
             x = paste(valine_col, "(IS Peak Area)"), y = "Extraction Buffer") + theme_minimal()
      
      plotdata_combined[plotdata_combined==0] <- 1
      cor_matrix <- plotdata_combined %>%
        dplyr::select(TIC, SampleMedian, ValineIS) %>% as.data.frame() %>%
        stats::cor(use = "complete.obs")
      
      cor_df <- reshape2::melt(cor_matrix)
      colnames(cor_df) <- c("Var1", "Var2", "Correlation")
      
      p4 <- ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
        geom_tile(color = "white") +
        geom_text(aes(label = round(Correlation, 2)), color = "black", size = 5) +
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
        theme_minimal() + labs(title = "Correlation Matrix") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
      
      p_all <- (p1 | p2 | p3 | p4)
      ggsave(file.path(SignalDrift_dir, paste0(add_to_name, "_Correlation_TIC_Median_ValineIS.svg")),
             plot = p_all, width = 24, height = 6)
      
      # Feature-wise correlations
      valine_vector <- Drift_data_ordered[, valine_col]
      valine_vector <- valine_vector[match(rownames(data), rownames(Drift_data_ordered))]
      
      feature_cor <- sapply(data, function(v) cor(v, valine_vector , use = "complete.obs"))
      feature_cor_df <- data.frame(Feature = names(feature_cor),
                                   CorrelationWithValineIS = feature_cor) %>%
        dplyr::arrange(CorrelationWithValineIS)
      
      top_neg <- head(feature_cor_df[order(feature_cor_df$CorrelationWithValineIS), ], 20) %>%
        dplyr::mutate(Feature = gsub("\\*", "", Feature))
      top_pos <- tail(feature_cor_df[order(feature_cor_df$CorrelationWithValineIS), ], 20) %>%
        dplyr::mutate(Feature = gsub("\\*", "", Feature))
      
      if (exists("ProjectFolder")) {
        rt_file <- file.path(ProjectFolder, "tsatsiki_v03_media/FINAL_TF_LIST.xlsx")
        if (file.exists(rt_file) && requireNamespace("readxl", quietly = TRUE)) {
          featRTs <- readxl::read_excel(rt_file)
          top_neg <- dplyr::left_join(top_neg, featRTs %>% dplyr::select(`Standardized name`, RT),
                                      by = c("Feature" = "Standardized name")) %>%
            dplyr::mutate(Feature = paste0(Feature, " (RT=", round(RT, 2), ")"))
          top_pos <- dplyr::left_join(top_pos, featRTs %>% dplyr::select(`Standardized name`, RT),
                                      by = c("Feature" = "Standardized name")) %>%
            dplyr::mutate(Feature = paste0(Feature, " (RT=", round(RT, 2), ")"))
        }
      }
      
      p_hist <- ggplot(feature_cor_df, aes(x = CorrelationWithValineIS)) +
        geom_histogram(bins = 50, fill = "steelblue", color = "white") +
        labs(title = paste("Feature-wise Correlation with", valine_col),
             x = "Pearson Correlation", y = "Feature Count") +
        theme_minimal()
      
      p_neg <- ggplot(top_neg, aes(x = reorder(Feature, CorrelationWithValineIS),
                                   y = CorrelationWithValineIS)) +
        geom_col(fill = "firebrick") + coord_flip() +
        labs(title = "Top 20 Most Negatively Correlated Features",
             x = "Feature", y = "Correlation") + theme_minimal()
      
      p_pos <- ggplot(top_pos, aes(x = reorder(Feature, CorrelationWithValineIS),
                                   y = CorrelationWithValineIS)) +
        geom_col(fill = "darkgreen") + coord_flip() +
        labs(title = "Top 20 Most Positively Correlated Features",
             x = "Feature", y = "Correlation") + theme_minimal()
      
      combined_plot <- p_hist / (p_neg | p_pos)
      ggsave(file.path(SignalDrift_dir, paste0(add_to_name, "_Feature_wise_Valine-d8_Correlations.svg")),
             plot = combined_plot, width = 18, height = 10)
      
      valine_done <- TRUE
    } else {
      valine_done <- FALSE
    }
    
    invisible(list(
      outdir = SignalDrift_dir,
      pool_label = pool_label,
      valine_found = valine_done
    ))
  }))
}
print("#0.6. Load helper functions!")
```

```{r}
# ───────────────────────────────────────────────
### HELPER FUNCTIONS: PCA + SIGNAL DRIFT PLOTS --


# PCA grid (pairwise PC plots + labeled version)
make_pca_grid <- function(data, SSF,
                          add_to_name   = NULL,
                          project_code  = ProjectCode,
                          color_col     = "Conditions",
                          fig_size_cm   = c(60, 40)) {

  suppressMessages(suppressWarnings({

    # ── Output folder ───────────────────────────
    PCA_dir <- file.path(output.dir, "PCA_plots")
    if (!dir.exists(PCA_dir)) dir.create(PCA_dir, recursive = TRUE)

    # ── Labels (remove leading "ProjectCode_" if present) ───────────────
    if (!is.null(project_code) && nzchar(project_code)) {
      SSF$Label <- sub(paste0("^", project_code, "_"), "", rownames(SSF))
    } else {
      SSF$Label <- rownames(SSF)
    }

    # ── PCA input (safe) ────────────────────────
    X <- data
    X[is.na(X)] <- 0

    # Drop zero-variance features
    nzv <- apply(X, 2, sd, na.rm = TRUE) > 0
    if (!all(nzv)) X <- X[, nzv, drop = FALSE]
    if (ncol(X) < 2) stop("Not enough non-zero-variance features for PCA.")

    # ── Run PCA ─────────────────────────────────
    pca_res  <- prcomp(X, scale. = TRUE)
    n_pc_use <- min(5, ncol(pca_res$x))

    # ── Palette from SSF (if Color column exists) ───────────────────────
    pal <- NULL
    if ("Color" %in% colnames(SSF) && color_col %in% colnames(SSF)) {
      pal_df <- unique(SSF[, c(color_col, "Color")])
      pal <- setNames(pal_df$Color, pal_df[[color_col]])
    }

    # ── Plot builder ────────────────────────────
    build_grid <- function(label_points = FALSE) {

      plots <- vector("list", n_pc_use * n_pc_use)
      k <- 1

      for (i in seq_len(n_pc_use)) {
        for (j in seq_len(n_pc_use)) {

          if (i != j) {

            p <- autoplot(
              pca_res, data = SSF, x = i, y = j,
              colour = color_col, size = 2, alpha = 0.8, loadings = FALSE
            ) +
              theme_classic() +
              ggtitle(paste("PC", i, "vs PC", j)) +
              geom_hline(yintercept = 0, color = "black", linewidth = 0.1) +
              geom_vline(xintercept = 0, color = "black", linewidth = 0.1)

            if (!is.null(pal)) p <- p + scale_color_manual(values = pal)
            if (label_points) {
              p <- p + ggrepel::geom_text_repel(aes(label = Label), size = 2.5)
            }

          } else {
            # diagonal cell: keep empty to make the grid readable
            p <- ggplot() + theme_void()
          }

          plots[[k]] <- p
          k <- k + 1
        }
      }

      patchwork::wrap_plots(plots, ncol = n_pc_use)
    }

    # ── Build + save ────────────────────────────
    prefix <- if (!is.null(add_to_name) && nzchar(add_to_name)) paste0(add_to_name, "__") else ""

    p_plain   <- build_grid(FALSE)
    p_labeled <- build_grid(TRUE)

    ggsave(
      filename = file.path(PCA_dir, paste0(prefix, "PCA.svg")),
      plot = p_plain, width = fig_size_cm[1], height = fig_size_cm[2], units = "cm"
    )

    ggsave(
      filename = file.path(PCA_dir, paste0(prefix, "PCA_Labeled.svg")),
      plot = p_labeled, width = fig_size_cm[1], height = fig_size_cm[2], units = "cm"
    )

    invisible(list(plain = p_plain, labeled = p_labeled, pca = pca_res))
  }))
}


# Signal drift reports (median + TIC, with optional Valine-d8 diagnostics)
make_signal_drift_reports <- function(data, SSF,
                                      add_to_name     = NULL,
                                      project_code    = ProjectCode,
                                      valine_pattern  = valine_pattern_id) {

  # Silence warnings inside (without changing global state permanently)
  old_warn <- getOption("warn")
  options(warn = -1)
  on.exit(options(warn = old_warn), add = TRUE)

  suppressMessages(suppressWarnings({

    # ── Output folder ───────────────────────────
    SignalDrift_dir <- file.path(output.dir, "SignalDrift")
    if (!dir.exists(SignalDrift_dir)) dir.create(SignalDrift_dir, recursive = TRUE)

    # ── Palette from SSF ────────────────────────
    pal_df <- unique(SSF[, c("Conditions", "Color")])
    customColors <- setNames(pal_df$Color, pal_df$Conditions)

    pal_for <- function(groups) {
      used <- intersect(names(customColors), unique(as.character(groups)))
      customColors[used]
    }

    # ── Pool label (first condition containing "pool") ──────────────────
    v_cond <- as.character(SSF$Conditions)
    pool_label <- {
      u <- unique(v_cond[grepl("pool", tolower(v_cond))])
      if (length(u)) u[1] else NA_character_
    }

    # ── Order by RunOrder ───────────────────────
    valid_samples <- rownames(SSF)[order(SSF$RunOrder)]

    # ── Merge Conditions + data (keeps rownames) ────────────────────────
    Drift_data <- merge(SSF %>% dplyr::select(Conditions), data, by = 0, sort = FALSE)
    colnames(Drift_data)[1] <- "File.Name"
    rownames(Drift_data) <- Drift_data$File.Name

    # Enforce acquisition order
    Drift_data_ordered <- Drift_data[match(valid_samples, Drift_data$File.Name), , drop = FALSE]
    Drift_data_ordered <- Drift_data_ordered[!is.na(rownames(Drift_data_ordered)), ]
    Drift_data_ordered$File.Name <- NULL

    # Ensure numeric intensities
    Drift_data_ordered[, -1] <- sapply(Drift_data_ordered[, -1], as.numeric)

    # ── Compute drift metrics (Median + TIC) ────────────────────────────
    row_medians <- matrixStats::rowMedians(as.matrix(Drift_data_ordered[, -1]), na.rm = TRUE)
    row_sums    <- rowSums(as.matrix(Drift_data_ordered[, -1]), na.rm = TRUE)

    plotdata <- data.frame(
      SampleMedian = row_medians,
      TIC          = row_sums,
      SampleGroup  = Drift_data_ordered$Conditions,
      check.names  = FALSE
    ) %>% tibble::rownames_to_column("RowNames")

    plotdata$RowNames <- factor(plotdata$RowNames, levels = plotdata$RowNames)
    plotdata$Order    <- seq_len(nrow(plotdata))

    # filename prefix
    prefix <- if (!is.null(add_to_name) && nzchar(add_to_name)) paste0(add_to_name, "__") else ""

    # ── Plot helper ────────────────────────────
    make_drift_plot <- function(df, yvar, ylab, title, filename) {

      pal_use <- pal_for(df$SampleGroup)
      df$SampleGroup <- factor(as.character(df$SampleGroup), levels = names(pal_use))

      pool_df <- if (!is.na(pool_label)) {
        df[as.character(df$SampleGroup) == pool_label, , drop = FALSE]
      } else {
        NULL
      }

      p <- ggplot(df, aes(x = RowNames, y = .data[[yvar]], fill = SampleGroup)) +
        geom_col(position = "dodge") +
        geom_smooth(
          data = pool_df,
          aes(x = Order, y = .data[[yvar]]),
          method = "lm", se = FALSE,
          color = "red", inherit.aes = FALSE
        ) +
        scale_fill_manual(values = pal_use, drop = FALSE) +
        labs(title = title, x = "Samples by acquisition order", y = ylab) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 40, hjust = 1),
          plot.title  = element_text(size = 22)
        )

      ggsave(
        filename = file.path(SignalDrift_dir, paste0(prefix, filename)),
        plot = p, width = 10, height = 8
      )
    }

    # ── Core drift plots ────────────────────────
    make_drift_plot(plotdata, "SampleMedian", "Sample median",
                    "Signal drift: sample median", "SignalDrift_Median_SampleGroup.svg")
    make_drift_plot(plotdata, "TIC", "Total ion count (TIC)",
                    "Signal drift: TIC", "SignalDrift_TIC_SampleGroup.svg")

    # Pool vs Sample (binary view)
    plotdata_pool <- transform(
      plotdata,
      SampleGroup = if (!is.na(pool_label)) ifelse(SampleGroup == pool_label, pool_label, "Sample") else "Sample"
    )

    make_drift_plot(plotdata_pool, "SampleMedian", "Sample median",
                    "Signal drift: median (pool highlighted)", "SignalDrift_Median_PoolSamples.svg")
    make_drift_plot(plotdata_pool, "TIC", "Total ion count (TIC)",
                    "Signal drift: TIC (pool highlighted)", "SignalDrift_TIC_PoolSamples.svg")

    # ── Optional: Valine-d8 diagnostics ─────────
    valine_idx  <- grep(valine_pattern, tolower(colnames(Drift_data_ordered)), ignore.case = TRUE)
    valine_done <- FALSE

    if (length(valine_idx) > 0) {

      valine_col <- colnames(Drift_data_ordered)[valine_idx[1]]

      plotdataIS <- data.frame(
        ValineIS    = Drift_data_ordered[[valine_col]],
        SampleGroup = Drift_data_ordered$Conditions,
        RowNames    = rownames(Drift_data_ordered),
        check.names = FALSE
      )

      plotdataIS$Index   <- seq_len(nrow(plotdataIS))
      plotdataIS$RowNames <- factor(plotdataIS$RowNames,
                                    levels = rownames(Drift_data_ordered), ordered = TRUE)

      pal_use <- pal_for(plotdataIS$SampleGroup)

      val_cv <- function(x) round(100 * sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE), 2)
      ValSampCV <- val_cv(plotdataIS$ValineIS)
      ValPoolCV <- if (!is.na(pool_label)) val_cv(plotdataIS$ValineIS[plotdataIS$SampleGroup == pool_label]) else NA_real_

      # Valine drift plot
      p_val <- ggplot(plotdataIS, aes(x = RowNames, y = ValineIS, fill = SampleGroup)) +
        geom_col(position = "dodge") +
        geom_smooth(aes(x = Index, y = ValineIS),
                    method = "lm", se = FALSE, color = "red", inherit.aes = FALSE) +
        scale_y_continuous(limits = c(0, NA)) +
        scale_fill_manual(values = pal_use, drop = FALSE) +
        labs(
          title    = paste0("Signal drift: ", valine_col, " (internal standard)"),
          subtitle = paste("PoolCV =", ValPoolCV, "| SampleCV =", ValSampCV),
          x = "Samples by acquisition order", y = "Peak area"
        ) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 40, hjust = 1),
              plot.title  = element_text(size = 22))

      ggsave(
        filename = file.path(SignalDrift_dir, paste0(prefix, "SignalDrift_Valine-d8(IS).svg")),
        plot = p_val, width = 12, height = 10
      )

      # Combine metrics for correlations
      plotdata_combined <- merge(plotdata, plotdataIS %>% dplyr::select(-SampleGroup), by = "RowNames")

      cor_TIC_valine    <- cor(plotdata_combined$TIC,          plotdata_combined$ValineIS, use = "complete.obs")
      cor_median_valine <- cor(plotdata_combined$SampleMedian, plotdata_combined$ValineIS, use = "complete.obs")

      # Optional: Extraction Buffer vs IS
      eb_col <- grep("^Extraction\\s*Buffer$", names(SSF), ignore.case = TRUE, value = TRUE)

      if (length(eb_col) == 1) {
        plotdata_combined$ExtractionBuffer <- SSF[match(plotdata_combined$RowNames, rownames(SSF)), eb_col]
        plotdata_combined$ExtractionBuffer <- as.numeric(plotdata_combined$ExtractionBuffer)
        plotdata_combined$ExtractionBuffer[is.na(plotdata_combined$ExtractionBuffer)] <-
          mean(plotdata_combined$ExtractionBuffer, na.rm = TRUE)

        ok <- complete.cases(plotdata_combined$ExtractionBuffer, plotdata_combined$ValineIS)
        cor_ExtrBuffer_valine <- if (sum(ok) >= 2) {
          cor(plotdata_combined$ExtractionBuffer[ok], plotdata_combined$ValineIS[ok])
        } else {
          NA_real_
        }
      } else {
        plotdata_combined$ExtractionBuffer <- NA_real_
        cor_ExtrBuffer_valine <- NA_real_
      }

      # Scatter plots
      p1 <- ggplot(plotdata_combined, aes(x = ValineIS, y = TIC, color = SampleGroup)) +
        geom_point(size = 3) +
        geom_smooth(method = "lm", se = FALSE, color = "black") +
        scale_color_manual(values = pal_for(plotdata_combined$SampleGroup), drop = FALSE) +
        labs(
          title = paste("TIC vs", valine_col, "(r =", round(cor_TIC_valine, 3), ")"),
          x = paste(valine_col, "(IS peak area)"), y = "TIC"
        ) +
        theme_minimal()

      p2 <- ggplot(plotdata_combined, aes(x = ValineIS, y = SampleMedian, color = SampleGroup)) +
        geom_point(size = 3) +
        geom_smooth(method = "lm", se = FALSE, color = "black") +
        scale_color_manual(values = pal_for(plotdata_combined$SampleGroup), drop = FALSE) +
        labs(
          title = paste("Median vs", valine_col, "(r =", round(cor_median_valine, 3), ")"),
          x = paste(valine_col, "(IS peak area)"), y = "Sample median"
        ) +
        theme_minimal()

      p3 <- ggplot(plotdata_combined, aes(x = ValineIS, y = ExtractionBuffer, color = SampleGroup)) +
        geom_point(size = 3) +
        geom_smooth(method = "lm", se = FALSE, color = "black") +
        scale_color_manual(values = pal_for(plotdata_combined$SampleGroup), drop = FALSE) +
        labs(
          title = paste("Extraction buffer vs", valine_col, "(r =", round(cor_ExtrBuffer_valine, 3), ")"),
          x = paste(valine_col, "(IS peak area)"), y = "Extraction buffer"
        ) +
        theme_minimal()

      # Correlation heatmap (TIC / Median / IS)
      cor_matrix <- plotdata_combined %>%
        dplyr::select(TIC, SampleMedian, ValineIS) %>%
        as.data.frame() %>%
        stats::cor(use = "complete.obs")

      cor_df <- reshape2::melt(cor_matrix)
      colnames(cor_df) <- c("Var1", "Var2", "Correlation")

      p4 <- ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
        geom_tile(color = "white") +
        geom_text(aes(label = round(Correlation, 2)), color = "black", size = 5) +
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
        theme_minimal() +
        labs(title = "Correlation matrix") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

      p_all <- (p1 | p2 | p3 | p4)

      ggsave(
        filename = file.path(SignalDrift_dir, paste0(prefix, "Correlation_TIC_Median_ValineIS.svg")),
        plot = p_all, width = 24, height = 6
      )

      # Feature-wise correlations vs IS
      valine_vector <- Drift_data_ordered[, valine_col]
      valine_vector <- valine_vector[match(rownames(data), rownames(Drift_data_ordered))]

      feature_cor <- sapply(data, function(v) cor(v, valine_vector, use = "complete.obs"))
      feature_cor_df <- data.frame(
        Feature = names(feature_cor),
        CorrelationWithValineIS = feature_cor,
        check.names = FALSE
      ) %>% dplyr::arrange(CorrelationWithValineIS)

      top_neg <- head(feature_cor_df[order(feature_cor_df$CorrelationWithValineIS), ], 20) %>%
        dplyr::mutate(Feature = gsub("\\*", "", Feature))
      top_pos <- tail(feature_cor_df[order(feature_cor_df$CorrelationWithValineIS), ], 20) %>%
        dplyr::mutate(Feature = gsub("\\*", "", Feature))

      # Optional: annotate with RT if TF list exists
      if (exists("ProjectFolder")) {
        rt_file <- file.path(ProjectFolder, "tsatsiki_v03_media/FINAL_TF_LIST.xlsx")
        if (file.exists(rt_file) && requireNamespace("readxl", quietly = TRUE)) {
          featRTs <- readxl::read_excel(rt_file)

          top_neg <- dplyr::left_join(top_neg, featRTs %>% dplyr::select(`Standardized name`, RT),
                                      by = c("Feature" = "Standardized name")) %>%
            dplyr::mutate(Feature = paste0(Feature, " (RT=", round(RT, 2), ")"))

          top_pos <- dplyr::left_join(top_pos, featRTs %>% dplyr::select(`Standardized name`, RT),
                                      by = c("Feature" = "Standardized name")) %>%
            dplyr::mutate(Feature = paste0(Feature, " (RT=", round(RT, 2), ")"))
        }
      }

      p_hist <- ggplot(feature_cor_df, aes(x = CorrelationWithValineIS)) +
        geom_histogram(bins = 50, fill = "steelblue", color = "white") +
        labs(title = paste("Feature-wise correlation with", valine_col),
             x = "Pearson correlation", y = "Number of features") +
        theme_minimal()

      p_neg <- ggplot(top_neg, aes(x = reorder(Feature, CorrelationWithValineIS), y = CorrelationWithValineIS)) +
        geom_col(fill = "firebrick") +
        coord_flip() +
        labs(title = "Top 20 most negatively correlated features",
             x = "Feature", y = "Correlation") +
        theme_minimal()

      p_pos <- ggplot(top_pos, aes(x = reorder(Feature, CorrelationWithValineIS), y = CorrelationWithValineIS)) +
        geom_col(fill = "darkgreen") +
        coord_flip() +
        labs(title = "Top 20 most positively correlated features",
             x = "Feature", y = "Correlation") +
        theme_minimal()

      combined_plot <- p_hist / (p_neg | p_pos)

      ggsave(
        filename = file.path(SignalDrift_dir, paste0(prefix, "Feature_wise_Valine-d8_Correlations.svg")),
        plot = combined_plot, width = 18, height = 10
      )

      valine_done <- TRUE
    }

    invisible(list(
      outdir        = SignalDrift_dir,
      pool_label    = pool_label,
      valine_found  = valine_done
    ))
  }))
}


# ──────────────────────────────────────────────
# Output message
message("#0.6. Load helper functions — completed successfully.")

```

### ---- 1. Data Cleaning ---- 
# 1.0 Preparation.
```{r}
# Set sub folders
Data_cleaning_dir <- file.path(output.dir, "1. Data Cleaning")
if (!dir.exists(Data_cleaning_dir)) dir.create(Data_cleaning_dir, recursive = TRUE)

Metabolite_Metadata <- data.frame(matrix(ncol = 0, nrow = length(colnames(data))))
Metabolite_Metadata$Metabolite <- colnames(data)
# Metabolite_Metadata$`SNratio<3` <- rownames(Metabolite_Metadata) %in% gsub("[^A-Za-z0-9 ]", "", names(blank_dominated_features[blank_dominated_features]))
# Data_cleaning_dir
# , ProjectCode,"_",Sys.Date(), 

# Prepare datasets
if(CoRe==T){
  # Split the data into samples and Ctrl samples
  # for fun check
  PCA_SSF_samples <- SSF[media_sample_names,]
  PCA_data_samples <- data[media_sample_names,]
  
  # for fun check
  PCA_SSF_ctrl <- SSF[media_control_names,]
  PCA_data_ctrl <- data[media_control_names,]
  
  PCA_SSF_list <- list(PCA_SSF_samples,PCA_SSF_ctrl)
  PCA_data_list <- list(PCA_data_samples,PCA_data_ctrl)
  

}else{
  PCA_SSF_list <- list(SSF)
  PCA_data_list <- list(data)
}

res <- make_pca_grid(data, SSF, add_to_name = "1. Raw_Raw")
make_signal_drift_reports(data, SSF, add_to_name = "1. Raw_Raw")

if(CoRe==T){
  # Split the data into samples and Ctrl samples
  # for fun check
  SSF_samples <- SSF[media_sample_names,]
  data_samples <-  data[media_sample_names,]
  
  # for fun check
  SSF_ctrl <- SSF[media_control_names,]
  data_ctrl <- data[media_control_names,]
  
  SSF_list <- list(SSF_samples,SSF_ctrl)
  data_list <- list(data_samples,data_ctrl)
}else{
  SSF_list <- list(SSF)
  data_list <- list(data)
}
paste('### ---- 1. Data Cleaning - #1.0 Preparation Finished!')
```

#1.1. Incosistent Features for TF peak verification.
```{r}
# Start new col in metabolite metadata
Metabolite_Metadata$Inconsistent_peaks <- FALSE

# Loop over datasets
for(i in 1:length(SSF_list)){ #i=1
  
  # Dataset selection
  SSF = SSF_list[[i]]
  data = data_list[[i]]
  if(i==1){
    outname<- "sample_"
  }else{
    outname<- "ctrl_"
  }
  
  # Remove blanks from data
  unconcistentTFpeak_data <- data[!rownames(data) %in% rownames(SSF)[tolower(SSF$Conditions) == blank_samp_id],]
  unconcistentTFpeak_SSF <-  SSF[!rownames(SSF) %in% rownames(SSF)[tolower(SSF$Conditions) == blank_samp_id],]
  # Find columns with zeros
  cols_with_zeros <- colnames(unconcistentTFpeak_data)[colSums(unconcistentTFpeak_data == 0, na.rm = TRUE) > 0]
  
  # Get columns with at least one zero and subset the data
  unconcistentTFpeak_data_with_zeros <- unconcistentTFpeak_data[, cols_with_zeros]
  
  # Initialize result list
  result <- list()
  all_values_mixgroups <- vector()
  
  # Loop over metabolites with at least one zero
  for (col in colnames(unconcistentTFpeak_data_with_zeros)) {    # col = colnames(unconcistentTFpeak_data_with_zeros)[1]
    metabolite_values <- unconcistentTFpeak_data_with_zeros[[col]]
    groups <- unconcistentTFpeak_SSF$Conditions
    unique_groups <- unique(groups)
    group_results <- setNames(rep(FALSE, length(unique_groups)), unique_groups)
    
    for (grp in unique_groups) {   # grp = unique_groups[2]
      group_values <- metabolite_values[groups == grp]
      has_zero <- any(group_values == 0, na.rm = TRUE)
      has_value <- any(group_values != 0, na.rm = TRUE)
      group_results[grp] <- has_zero & has_value
      if (has_zero & has_value) {
        all_values_mixgroups <- c(all_values_mixgroups, group_values)
      }
    }
    result[[col]] <- group_results
  }
  
  # Build result table
  result_df <- do.call(rbind, result)
  result_df_filtered <- result_df[apply(result_df, 1, any), ]  # Keep features with at least one TRUE
  result_df_filtered <- as.data.frame(t(result_df_filtered))
  result_df_filtered <- rownames_to_column(result_df_filtered, "SampleGroup")
  
  # Get corresponding values
  result_df_filtered_values <- unconcistentTFpeak_data[, colnames(unconcistentTFpeak_data) %in% colnames(result_df_filtered)]
  result_df_filtered_values <- rownames_to_column(result_df_filtered_values, "SampleGroup")
  result_df_filtered_values$Sample <- result_df_filtered_values$SampleGroup
  result_df_filtered_values$SampleGroup <- unconcistentTFpeak_SSF[result_df_filtered_values$SampleGroup, "Conditions"]
  result_df_filtered_values <- result_df_filtered_values %>% relocate(Sample, .before = everything())
  
  # add cross to the names in the original data, Example: add † to each column name
  Incosistent_metab_names <- colnames(data)[colnames(data) %in% colnames(result_df_filtered)]
  colnames(data_raw)[gsub("\u2020", "",colnames(data_raw)) %in% colnames(result_df_filtered)] <- paste0(Incosistent_metab_names, "\u2020")  
  # "\u2020" = †
  
  # Excel export
  wb <- createWorkbook()
  
  # Sheet 1: Boolean matrix
  addWorksheet(wb, "boolean")
  writeData(wb, "boolean", result_df_filtered)
  conditionalFormatting(wb, "boolean",
                        cols = 2:ncol(result_df_filtered),
                        rows = 2:(nrow(result_df_filtered) + 1),
                        type = "contains",
                        rule = "TRUE",
                        style = createStyle(bgFill = "#FFC7CE")
  )
  
  # Sheet 2: Raw intensity values
  addWorksheet(wb, "values")
  writeData(wb, "values", result_df_filtered_values)
  
  # Style for highlighting
  redStyle <- createStyle(fgFill = "#FFC7CE", fontColour = "#9C0006")
  
  # Highlight cells where inconsistency was detected
  for (r in 1:nrow(result_df_filtered)) {
    sample_group <- result_df_filtered[r, "SampleGroup"]
    for (c in colnames(result_df_filtered)[-1]) {
      if (isTRUE(result_df_filtered[r, c])) {
        matching_rows <- which(result_df_filtered_values$SampleGroup == sample_group)
        matching_col <- which(colnames(result_df_filtered_values) == c)
        for (row in matching_rows) {
          addStyle(wb, "values", style = redStyle, rows = row + 1, cols = matching_col, stack = TRUE)
        }
      }
    }
  }
  
  # Save the Excel file
  saveWorkbook(wb, file = file.path(Data_cleaning_dir, paste0("Incosistent_Features_for_TF_peak_verification_",outname ,ProjectCode,"_",Sys.Date(), ".xlsx")), overwrite = TRUE)
  
  
  
  
  ### ----- Make data values distribution plots with highlighted the values of Inconsistent peaks. ---- ###
  
  # Flatten the data into one long vector
  all_values <- as.vector(as.matrix(unconcistentTFpeak_data))
  
  # Create a dataframe marking whether each value is in your vector
  plot_df <- data.frame(
    Value = all_values,
    Highlight = ifelse(all_values %in% all_values_mixgroups, "Highlighted", "Background"))
  
  # Plot with x-axis limited from 0 to 100,000
  mixedGroupDist <- ggplot(plot_df, aes(x = Value, fill = Highlight)) +
    geom_histogram(bins = 300000, alpha = 0.7, position = "identity") +
    scale_fill_manual(values = c("Background" = "grey70", "Highlighted" = "red")) +
    coord_cartesian(xlim = c(0, 20000000)) +
    labs(title = "Distribution of All Values with Highlighted values of Incosistent Features",
         x = "Value",
         y = "Count") +
    theme_minimal()
  
  # Save plot
  ggsave(filename = file.path(Data_cleaning_dir, paste0("zeroMixedGroups_Distribution_",outname , ProjectCode,"_",Sys.Date(), ".svg")), 
         plot = mixedGroupDist, width = 14, height = 8)
  
  # Update Metabolite metadata
  Metabolite_Metadata$Inconsistent_peaks <-  Metabolite_Metadata$Inconsistent_peaks |Metabolite_Metadata$Metabolite %in% Incosistent_metab_names
  
  write.xlsx(Metabolite_Metadata, file.path(Data_cleaning_dir,"Metabolite_Metadata.xlsx"), colNames = TRUE)
}
paste0("#1.1. Incosistent Features for TF peak verification Finished!")
```

#1.2. Remove metabolites in blanks.
```{r}
Metabolite_Metadata$`SNratio<3`  <- FALSE
for(i in 1:length(SSF_list)){ #i=1
  
  SSF = SSF_list[[i]]
  data = data_list[[i]]
  PCA_SSF = PCA_SSF_list[[i]]
  PCA_data = PCA_data_list[[i]]
  if(i==1){
    outname<- "sample_"
  }else{
    outname<- "ctrl_"
  }
  
  # Split data into blanks and real samples
  blank_rows <- rownames(SSF)[grep("Blank", SSF$Conditions)]
  sample_rows <- rownames(SSF)[!rownames(SSF) %in% blank_rows]
  
  # Subset data
  data_blanks <- data[blank_rows, , drop = FALSE]
  data_samples <- data[sample_rows, , drop = FALSE]
  
  # Calculate feature means across blanks and samples
  blank_means <- colMeans(data_blanks, na.rm = TRUE)
  blank_means[is.na(blank_means)] <- 0
  
  sample_means <- colMeans(data_samples, na.rm = TRUE)
  sample_means[is.na(sample_means)] <- 0
  
  # Set Signal to Noise ratio threshold (e.g., sample mean must be >3x blank mean)
  SignaltoNoiseRatio_threshold <- SignaltoNoiseRatio_threshold
  keep_features <- sample_means > (SignaltoNoiseRatio_threshold * blank_means)
  
  # Keep only features passing the filter
  filtered_data <- data_samples[, keep_features, drop = FALSE]
  
  # For features not passing the filter, set their values in samples to zero
  blank_dominated_features <- !keep_features
  blank_dominated_data <- data[, blank_dominated_features, drop = FALSE]
  
  # Make barplot of Blank dominated metabolites and save it for visual inspection.
  blank_drift <- list()
  for (metab in colnames(blank_dominated_data)) {   # metab = colnames(data)[1]
    # Merge metabolite values with metadata
    plotdata <- merge(blank_dominated_data %>% select(all_of(metab)), SSF, by = 0)
    
    # Ensure RunOrder is numeric and sort
    plotdata <- plotdata[order(plotdata$RunOrder), ]
    
    # Make Row.names a factor ordered by RunOrder
    plotdata$Row.names <- factor(plotdata$Row.names, levels = plotdata$Row.names)
    
    # Plot with Colors column used for custom colors but Conditions for legend
    p <- ggplot(plotdata, aes(x = Row.names, y = .data[[metab]], fill = Conditions)) +
      geom_col(position = "dodge") +
      scale_fill_manual(values = setNames(plotdata$Color, plotdata$Conditions)) + # map colors
      labs(title = paste(metab),
           x = "Samples (Run Order)", y = "Intensity") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),  # rotate x labels
            plot.title = element_text(size = 14))
    
    # Store plot
    blank_drift[[metab]] <- p
  }
  
  # Save blank Bar-plots
  BlankPlotsloc <- file.path(Data_cleaning_dir,paste0("BlankBarPlots_" ,outname , ProjectCode,"_",Sys.Date(), ".pdf"))
  pdf(BlankPlotsloc, width = 8, height = 8) # Adjust width and height as needed
  
  # Iterate over the list of plots and print each one
  for (plot in blank_drift) {
    print(plot)
  }
  
  # Close the PDF device
  dev.off()
  
  # # remove
  # data <- filtered_data
  # SSF <- SSF[rownames(SSF) %in% rownames(data),]
  
  # Update metabolite metadata
  Metabolite_Metadata$`SNratio<3` <-  Metabolite_Metadata$`SNratio<3` |Metabolite_Metadata$Metabolite %in% gsub("\u2020", "", names(blank_dominated_features[blank_dominated_features]))
  
  write.xlsx(Metabolite_Metadata, file.path(Data_cleaning_dir,"Metabolite_Metadata.xlsx"), colNames = TRUE)
  
  # Remove blanks
  
  SSF <- SSF[-grep(blank_samp_id , rownames(SSF)),]
  data <- data[-grep(blank_samp_id , rownames(data)),]  
  # Remove blanks from data
  SSF_list[[i]] <- SSF
  data_list[[i]] <- data
  
  
  # PCA
  # # remove
  PCA_data <- filtered_data
  PCA_SSF <- SSF[rownames(SSF) %in% rownames(PCA_data),]
  make_pca_grid(PCA_data, PCA_SSF, add_to_name = paste0("2. ",outname,"Raw_Blank_removed"))
  make_signal_drift_reports(PCA_data, PCA_SSF, add_to_name =  paste0("2. ",outname,"Raw_Blank_removed"))
  PCA_SSF_list[[i]] <- PCA_SSF
  PCA_data_list[[i]] <- PCA_data
}
paste0("#1.2. Remove metabolites in blanks Finished!")
```


#1.3. Feature filtering for missing values in samples and pools. 80% modified rule.
```{r}
#Here we use the 80% modified rule so we allow for ~20% missing values. We look at the smallest sample group. Fοr the pools we do the same but we always have to have at least 4 Pool samples.

Metabolite_Metadata$FeatureFiltering  <- FALSE
for(i in 1:length(SSF_list)){ #i=1
  
  SSF = SSF_list[[i]]
  data = data_list[[i]]
  PCA_SSF = PCA_SSF_list[[i]]
  PCA_data = PCA_data_list[[i]]
  if(i==1){
    outname<- "sample_"
  }else{
    outname<- "ctrl_"
  }
  
  # remake the zeros into NAs 
  data[data==0] <- NA
  # Parameters
  Feature_Filt_Value <- Feature_Filt_Value # Set sample filtering value
  Feature_Filt_Value_Pools <- Feature_Filt_Value_Pools # Set pool filtering value \
  # If we have 5 samples/pools, we allow 1 0/NA in 5 which means 75 not 80
  # If we have 4 samples/pools, we allow 1 0/NA in 4 which means 75 not 80
  
  # Get features to filter for each group
  filter_results <- list()
  
  ## Here I am excluding pool ###
  for (group in setdiff(unique(SSF$Conditions), c("Blank")) ) { # group = setdiff(unique(SSF$Conditions), "Blank")[1]
    
    threshold <- if (isTRUE(group == SSF$Conditions[grep("pool", tolower(SSF$Conditions))[1]]))
  Feature_Filt_Value_Pools else Feature_Filt_Value
    
    # Subset data and conditions
    idx <- which(SSF$Conditions == group)
    sub_data <- data[idx, , drop = FALSE]
    sub_conditions <- SSF$Conditions[idx]
    
    # Skip if no data
    if (length(idx) == 0 || length(unique(sub_conditions)) < 1) {
      filter_results[[group]] <- character(0)
      next
    }
    
    # Split by condition
    split_data <- split(sub_data, sub_conditions)
    #split_data <- split(data, SSF$Conditions)
    
    feature_missing <- c() 
    
    
    for (df in split_data) { # df = split_data[1]
      df <- as.data.frame(df)
      for (col in seq_len(ncol(df))) {    #   col=18
        if (sum(is.na(df[, col])) > round((1 - threshold) * nrow(df)) ) { # 0.8 means, we allow up but NOT equal to 20 NA vales.
          # We Use greater than the threshold. So, 1 NA in 5 non NA is allowed.
          feature_missing <- c(feature_missing, col)
        }
      }
    }
    filter_results[[group]] <-  colnames(data)[unique(feature_missing)]
  }
  
  # Remove a feature if it is not found in the pools
  to_filter <- c()
  to_filter <- filter_results[[ as.character(SSF$Conditions[ grep("pool", tolower(SSF$Conditions))[1] ]) ]]
  
  # Keep only features missing in *all* conditions ignoring pools
  feature_missing <- unlist(filter_results)
  if (length(feature_missing) > 0) {
    count <- table(feature_missing)
    for (c in names(count)) { 
      if (count[[c]] == length(unique(SSF$Conditions[-grep("pool", tolower(SSF$Conditions))])) ) {
        to_filter <- c(to_filter, c)
      }
    }
  }
  
  # Combine
  filtered_features <- unique(to_filter)
  
  # Filter data now. Otherwise they are removed when its time for them to be removed. 
  #data <- data[, !colnames(data) %in% filtered_features, drop = FALSE]
  
  features_filtered_data <- data[,filtered_features]
  features_filtered_data <- rownames_to_column(features_filtered_data, "Samples")
  
  # Save values for visual inspection
  write.xlsx(features_filtered_data, file.path(Data_cleaning_dir,"Features_FIltered_data.csv"), colNames = TRUE)
  
  # Update metabolite metadata
  Metabolite_Metadata$FeatureFiltering <-  Metabolite_Metadata$FeatureFiltering |Metabolite_Metadata$Metabolite %in% gsub("\u2020", "",filtered_features)
  
  write.xlsx(Metabolite_Metadata, file.path(Data_cleaning_dir,"Metabolite_Metadata.xlsx"), colNames = TRUE)
  
  # PCA
  # # remove
  PCA_data <- PCA_data <- PCA_data[, !colnames(PCA_data) %in% filtered_features, drop = FALSE]
  PCA_SSF <- SSF[rownames(SSF) %in% rownames(PCA_data),]
  make_pca_grid(PCA_data, PCA_SSF, add_to_name =  paste0("3. ",outname,"NA_filtered"))
  make_signal_drift_reports(PCA_data, PCA_SSF, add_to_name = paste0("3. ",outname,"NA_filtered"))
  PCA_SSF_list[[i]] <- PCA_SSF
  PCA_data_list[[i]] <- PCA_data
}
paste0("#1.3. Feature filtering for missing values in samples and pools. 80% modified rule Finished!")
```

#1.4. CV filtering. 30% filtering.
```{r}
Threshold_cv = Threshold_cv

for(i in 1:length(SSF_list)){ # i=1
  
  SSF = SSF_list[[i]]
  data = data_list[[i]]
  PCA_SSF = PCA_SSF_list[[i]]
  PCA_data = PCA_data_list[[i]]
  if(i==1){
    outname<- "sample_"
  }else{
    outname<- "ctrl_"
  }
  
   if(CoRe == T & i==1 ){
   ## -------- Pool Coefficient of variation Estimation ------ ##
  #A. Pool
  pool_data <- data[rownames(SSF)[SSF$Conditions == SSF$Conditions[grep("pool", tolower(SSF$Conditions))[1]]], ]
  # Calculate the Coefficient of Variation
  result_df <- apply(pool_data, 2,  function(x) { (sd(x, na.rm =T)/  mean(x, na.rm =T))*100 }  ) %>% t()%>% as.data.frame()
  rownames(result_df)[1] <- "Pool_CV"
  # reorder 
  result_df <- result_df[, names(pool_data)]
  # calculate the percentage of  NAs
  NAvector <- apply(pool_data, 2,  function(x) { (sum(is.na(x))/length(x))*100 }  )#%>% t()%>% as.data.frame()
  # Make the final dataframe
  result_df_final <- result_df %>% t() %>% as.data.frame()%>% mutate(HighCV = Pool_CV > Threshold_cv)# %>% as.data.frame()
  result_df_final$Pool_MissingValuePercentage <- NAvector
  rownames(result_df_final)<- colnames(pool_data)
  
  #b. Sample
  sample_data <- data[rownames(SSF)[SSF$Conditions != SSF$Conditions[grep("pool", tolower(SSF$Conditions))[1]]], ]
  # Calculate the Coefficient of Variation
  result_df <- apply(sample_data, 2,  function(x) { (sd(x, na.rm =T)/  mean(x, na.rm =T))*100 }  ) %>% t()%>% as.data.frame()
  rownames(result_df)[1] <- "Sample_CV"
  # reorder 
  result_df <- result_df[, names(sample_data)]
  # calculate the percentage of  NAs
  NAvector <- apply(sample_data, 2,  function(x) { (sum(is.na(x))/length(x))*100 }  )#%>% t()%>% as.data.frame()
  # Make the final dataframe
  result_df_final <- merge(result_df_final, result_df %>% t() %>% as.data.frame(), by=0)
  colnames(result_df_final)[1] <- "Metabolite" 
  result_df_final <- result_df_final[match(colnames(data), result_df_final$Metabolite), ]
  
  #### -------------------------- CV histogram --------------------- ####
  # Reshape to long format
  long_df <- result_df_final %>%
    select(Metabolite, Pool_CV, Sample_CV) %>%
    pivot_longer(cols = c(Pool_CV, Sample_CV), names_to = "Type", values_to = "CV") %>%
    mutate(Type = ifelse(Type == "Pool_CV", "Pool", "Sample"))
  
  # Create plot with stronger color distinction
  HistCVcombined <- ggplot(long_df, aes(x = CV, fill = Type, color = Type)) +
    geom_density(alpha = 0.3, size = 1.0) +
    geom_vline(aes(xintercept = Threshold_cv), color = "black", linetype = "dashed", size = 1) +
    labs(title = "Coefficient of Variation for Pool and Sample",
         x = "Coefficient of Variation (CV%)",
         y = "Density",
         caption = paste0(nrow(result_df_final), " total features")) +
    scale_fill_manual(values = c("Pool" = "red", "Sample" = "blue")) +
    scale_color_manual(values = c("Pool" = "red", "Sample" = "blue")) +
    theme_classic()
  
  # Add Valine-d8 (IS) line if present
  if (length(grep(valine_pattern_id, tolower(result_df_final$Metabolite))) > 0) {
    val_cv_pool <- result_df_final$Pool_CV[grep(valine_pattern_id, tolower(result_df_final$Metabolite))]
    val_cv_samp <- result_df_final$Sample_CV[grep(valine_pattern_id, tolower(result_df_final$Metabolite))]
    
    HistCVcombined <- HistCVcombined +
      geom_vline(xintercept = val_cv_pool, color = "red", linetype = "dashed", size = 1) +
      geom_vline(xintercept = val_cv_samp, color = "blue", linetype = "dashed", size = 1) +
      scale_x_continuous(breaks = c(30, 50, 100, val_cv_pool),
                         labels = c("30", "50", "100", "Valine-d8 (IS)"))
  }
  
  # Save
  ggsave(filename = file.path(Data_cleaning_dir,paste0(outname ,"CV_Histogram.svg")), 
         plot = HistCVcombined, width = 12, height = 6)
  
  message("n of metabolites removed because of RSD > " ,Threshold_cv, "% : " , sum(result_df_final$HighCV==T, na.rm = T))
  
  # Remove high cv features
  # sample_data_high_cv_feats <- colnames(data)[result_df_final$HighCV==T]
  # data <- data[,!colnames(data) %in% sample_data_high_cv_feats]
  
  # Save resulting table
  result_df_final$Metabolite <- gsub("\u2020", "",result_df_final$Metabolite)
  
  colnames(result_df_final)[colnames(result_df_final) %in% c("HighCV", "Pool_CV", "Pool_MissingValuePercentage", "Sample_CV")] <- 
    paste0(outname, c("HighCV", "Pool_CV","Pool_MissingValuePercentage" ,"Sample_CV"))
  
  Metabolite_Metadata <- merge(Metabolite_Metadata, result_df_final, "Metabolite")
  write.xlsx(Metabolite_Metadata, file.path(Data_cleaning_dir,"Metabolite_Metadata.xlsx"), colNames = TRUE)
   }
  
     if(CoRe == T & i==2 ){
       Metabolite_Metadata <- read.xlsx( file.path(Data_cleaning_dir, "Metabolite_Metadata.xlsx"))
     }
 
  
  # PCA
  # # remove
  sample_data_high_cv_feats <- colnames(data)[Metabolite_Metadata$sample_Pool_CV ==T]
  PCA_data <- PCA_data[,!colnames(PCA_data) %in% sample_data_high_cv_feats]
  PCA_SSF <- SSF[rownames(SSF) %in% rownames(PCA_data),]
  make_pca_grid(PCA_data, PCA_SSF, add_to_name = paste0("4. ",outname,"CV_filtered"))
  make_signal_drift_reports(PCA_data, PCA_SSF, add_to_name =  paste0("4. ",outname,"CV_filtered"))
  PCA_SSF_list[[i]] <- PCA_SSF
  PCA_data_list[[i]] <- PCA_data
}

paste0("#1.4. CV filtering. 30% filtering Finished!")
```

#1.5. Single Outlier Values filtering. Skip generally
```{r eval = FALSE}
data$SampleID <- rownames(data)
SSF$SampleID <- rownames(SSF)

long_data <- data %>%
  pivot_longer(-SampleID, names_to = "Metabolite", values_to = "Value") %>%
  left_join(SSF[, c("SampleID", "Conditions")], by = "SampleID")

# Log data to make them ~ normally distributed
#long_data$Value <- log2(long_data$Value)
# Placeholder to collect outliers
outlier_results <- list()

for(selected_metabolite in  unique(long_data$Metabolite)){ # selected_metabolite = unique(long_data$Metabolite)[1]
  if (selected_metabolite == "Valine-d8") next
  # Filter for that metabolite
  metab_data <- long_data %>% filter(Metabolite == selected_metabolite)
  
  # Loop over conditions
  for (cond in unique(metab_data$Conditions)) { # cond = unique(metab_data$Conditions)[1]
    group_data <- metab_data %>% filter(Conditions == cond)
    
    if(sum(!is.na(group_data$Value))>1){
      for (i in seq_len(nrow(group_data))) {  #   i =  seq_len(nrow(group_data))[1]
        test_val <- group_data$Value[i]
        test_sample <- group_data$SampleID[i]
        
        # Leave-one-out: remove current value
        rest <- group_data$Value[-i]
        
        if(length(rest[!is.na(rest)])>1){
          m <- mean(rest, na.rm = TRUE)
          s <- sd(rest, na.rm = TRUE)
          
          lower <- m - 4 * s
          upper <- m + 4 * s
          
          if (!is.na(test_val) && (test_val < lower || test_val > upper)) {
            outlier_results[[length(outlier_results) + 1]] <- data.frame(
              SampleID = test_sample,
              Condition = cond,
              Value = test_val,
              Mean_rest = m,
              SD_rest = s,
              Metabolite = selected_metabolite
            )
          }  
        }
      }
    }
  }
}

# Combine and print
if (length(outlier_results) > 0) {
  # Make outliers values into a df
  outlier_results <- outlier_results[!sapply(outlier_results, is.null)]
  # Combine all into a single data.frame
  outlier_df <- do.call(rbind, outlier_results)
  # View result
  head(outlier_df)
  
  message(nrow(outlier_df)  ," Outlier values found.\n")
} else {
  cat("No outliers found.\n")
}

#remove the sample IDs
data$SampleID <- NULL
SSF$SampleID <- NULL

# remove features  skip

# # Loop through each outlier and set it to NA
# for (i in seq_len(nrow(outlier_df))) {  # i =seq_len(nrow(outlier_df))[1]
#   samp <- outlier_df$SampleID[i]
#   feat <- outlier_df$Metabolite[i]
#   data[samp, feat] <- NA
# }

# DO NOT remove single outler. Its not working properly yet. is gives hints but dont remove
# # measure cv. if its high then 
# x <- group_data$Value
# med <- median(x)
# mad_val <- mad(x)  # default is scaled for normal distribution
# 
# # Robust Z-scores
# robust_z <- (x - med) / mad_val
# which(abs(robust_z) > 2.5)
# if its found agaib then remove? or do something about
paste0("#1.5. Single Outlier Values filtering!")
```

Filter features

Here I am filtering only with Feature filtering. In standard pipeline we filter with the filtered column.
```{r}
# Ensure metabolite order from data colnames
met_order <-    gsub("\u2020", "", colnames(data_raw))

# reorder metadata to match that order
Metabolite_Metadata <- Metabolite_Metadata[match(met_order, Metabolite_Metadata$Metabolite), ]

# Metabolite_Metadata <- Metabolite_Metadata %>%
#   mutate(Filtered = if_any(
#     everything(),  # check all other columns
#     ~ .x == TRUE))

Metabolite_Metadata <- Metabolite_Metadata %>%
  mutate(Filtered = if_any(-Inconsistent_peaks, ~ .x == TRUE))

write.xlsx(Metabolite_Metadata, file.path(Data_cleaning_dir,"Metabolite_Metadata.xlsx"), colNames = TRUE)

# filter
metabs_to_remvove <- Metabolite_Metadata %>% filter(Filtered) %>% pull(Metabolite)

## Here we use the data_raw but the background subtracted data data_raw
data <- data_raw[, !gsub("†", "", colnames(data_raw)) %in% metabs_to_remvove, drop = FALSE]

# Remove blanks
blank_samples <- rownames(SSF_raw)[tolower(SSF_raw$Conditions) == "blank"]

data <- data[!rownames(data) %in% blank_samples, ]
SSF <- SSF_raw[!rownames(SSF_raw) %in% blank_samples, ]

paste0("### ---- 1. Data Cleaning ---- Finished!")
```



### ----- 2. Data Normalisation -----
#2.0 Preparation
```{r}
# Set sub folders
Data_normalisation_dir <- file.path(output.dir, "2. Data Normalisation")
if (!dir.exists(Data_normalisation_dir)) dir.create(Data_normalisation_dir, recursive = TRUE)

# Metabolite_Metadata$`SNratio<3` <- rownames(Metabolite_Metadata) %in% gsub("[^A-Za-z0-9 ]", "", names(blank_dominated_features[blank_dominated_features]))
# Data_cleaning_dir
# , ProjectCode,"_",Sys.Date(), 
  
if(CoRe==T){
  # Split the data into samples and Ctrl samples
  # for fun check
  SSF_samples <- SSF[media_sample_names[!grepl(blank_samp_id, media_sample_names)],]
  data_samples <- data[media_sample_names[!grepl(blank_samp_id, media_sample_names)],]
  
  # for fun check
  SSF_ctrl <- SSF[media_control_names[!grepl(blank_samp_id, media_control_names)],]
  data_ctrl <- data[media_control_names[!grepl(blank_samp_id, media_control_names)],]
  
  SSF_list <- list(SSF_samples,SSF_ctrl)
  data_list <- list(data_samples,data_ctrl)
}else{
  SSF_list <- list(SSF)
  data_list <- list(data)
}
paste0("#2.0 Preparation Finished!")
```

#2.1. IS normalisation valine-d8

# Check if TIC is required or VAlineIS normalisation
```{r}
for(i in 1:length(SSF_list)){ # i=1
  
  SSF = SSF_list[[i]]
  data = data_list[[i]]
  PCA_SSF = PCA_SSF_list[[i]]
  PCA_data = PCA_data_list[[i]]
  if(i==1){
    outname<- "sample_"
  }else{
    outname<- "ctrl_"
  }
  
  ### ### Check TIC -----------------
  SSF$TIC <- rowSums(data[rownames(SSF), , drop = FALSE], na.rm = TRUE)
  
  # 2) make a discrete key (because your function uses scale_color_manual)
  SSF$TIC_key <- factor(rank(-SSF$TIC, ties.method = "first"))  # 1 = highest TIC
  
  # 3) map TIC rank to a gradient (high=red, low=blue)
  tic_pal <- colorRampPalette(c("red","blue"))(nrow(SSF))
  SSF$TIC_hex <- tic_pal[as.integer(SSF$TIC_key)]
  
  # 4) feed your function: set Color to hex, color_col to TIC_key
  SSF$colorsave <- SSF$Color
  SSF$Color <- SSF$TIC_hex
  
  make_pca_grid(data, SSF,
                add_to_name = paste0("5.0 " ,outname, "TIC_gradient_CHECK"),
                color_col   = "TIC_key")
  
  # Fix SSF
  SSF$Color <- SSF$colorsave
  SSF$TIC_hex <- NULL
  SSF$TIC_key <- NULL
  SSF$colorsave <- NULL
  SSF$TIC <- NULL
  
  ### ### Check Valine -----------
  # 1) Valine intensity per sample
  SSF$VAL <- data[rownames(SSF), "valine-d8"]
  
  # 2) rank (highest = 1)
  SSF$VAL_key <- factor(rank(-SSF$VAL, ties.method = "first"))
  
  # 3) gradient (high=red, low=blue)
  val_pal <- colorRampPalette(c("red","blue"))(nrow(SSF))
  SSF$VAL_hex <- val_pal[as.integer(SSF$VAL_key)]
  
  # 4) feed PCA function
  SSF$colorsave <- SSF$Color
  SSF$Color <- SSF$VAL_hex
  
  make_pca_grid(data, SSF,
                add_to_name = paste0("5.1 " ,outname, "Valine_gradient_CHECK"),
                color_col   = "VAL_key")
  
  # restore SSF
  SSF$Color <- SSF$colorsave
  SSF$VAL_hex <- NULL
  SSF$VAL_key <- NULL
  SSF$colorsave <- NULL
  SSF$VAL <- NULL
}
```

# TIC normalisation
```{r }
# shoud you norlaise with tic? CHEKC IF YOU SEE THE PATTERN IN PCA

for(i in 1:length(SSF_list)){ # i=1
  
  SSF = SSF_list[[i]]
  data = data_list[[i]]
  PCA_SSF = PCA_SSF_list[[i]]
  PCA_data = PCA_data_list[[i]]
  if(i==1){
    outname<- "sample_"
  }else{
    outname<- "ctrl_"
  }
  
  tic <- rowSums(data_list[[i]], na.rm = TRUE)
  tic_ref <- median(tic, na.rm = TRUE)
  
  data_tic <- sweep(data_list[[i]], 1, tic, "/") * tic_ref
  
  SSF <- SSF_list[[i]]
  # PCA
  make_pca_grid(data_tic, SSF, add_to_name = paste0("5.5 " ,outname, "TIC_normalised"))
  make_signal_drift_reports(data_tic, SSF, add_to_name = paste0("5.5 " ,outname , "TIC_normalised"))
  
  make_pca_grid(data_tic, SSF, add_to_name = paste0("5.5 TIC_normalised_Big"), fig_size_cm = c(120, 80))
  
  data_list[[i]] <-data_tic 
}
paste0("#2.1 TIC normalisation Finished!")
```


#2.2. Signal Drift normalisation using LoessQC

```{r}
library(qcrlscR)

suppressPackageStartupMessages({
  library(qcrlscR)
  library(dplyr)
  library(tibble)
  library(ggplot2)
  library(gridExtra)
  library(openxlsx)
})

# ---------- INPUT ----------
data <- data_list[[1]]
SSF <- SSF_list[[1]]

# ---------- ORDER BY RUN ORDER ----------
ord <- order(SSF$RunOrder)
data_o <- data[ord, , drop = FALSE]
SSF_o  <- SSF[ord, , drop = FALSE]

all_samples <- rownames(SSF_o)
feat_names  <- colnames(data_o)

# QC/sample factor for qcrlscR
cls_qc <- ifelse(SSF_o$Conditions == "Pool", "qc", "sample")
cls_qc <- factor(cls_qc, levels = c("qc", "sample"))
cls_bl <- factor(rep(1, nrow(SSF_o)))  # single batch

# colors (expects SSF has: Conditions + Color)
pal_map <- unique(SSF_o[, c("Conditions", "Color")])
customColors <- setNames(pal_map$Color, pal_map$Conditions)

# QC samples (Pools)
qc_samples <- rownames(SSF_o)[tolower(SSF_o$Conditions) == "pool" | grepl("pool", tolower(SSF_o$Conditions))]
samp_ids   <- setdiff(all_samples, qc_samples)
qc_idx     <- which(all_samples %in% qc_samples)

# group labels (within-group CV) only for NON-QC samples
group_labels <- SSF_o$Conditions
names(group_labels) <- rownames(SSF_o)
group_labels_samp <- droplevels(factor(group_labels[samp_ids]))

# ---------- SAFE HELPERS ----------
cv_safe <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  m <- mean(x, na.rm = TRUE)
  if (!is.finite(m) || m == 0) return(NA_real_)
  100 * sd(x, na.rm = TRUE) / m
}

within_cv_safe <- function(x, groups) {
  x <- as.numeric(x)
  ok <- is.finite(x) & !is.na(groups)
  x <- x[ok]
  g <- droplevels(groups[ok])
  
  if (length(x) < 2) return(NA_real_)
  
  cvs <- tapply(x, g, function(v) {
    v <- v[is.finite(v)]
    if (length(v) < 2) return(NA_real_)
    m <- mean(v, na.rm = TRUE)
    if (!is.finite(m) || m == 0) return(NA_real_)
    100 * sd(v, na.rm = TRUE) / m
  })
  
  if (all(is.na(cvs))) NA_real_ else mean(cvs, na.rm = TRUE)
}

fmt1 <- function(x) ifelse(is.finite(x), sprintf("%.1f", x), "NA")

# Excel writer (3 decimals formatting)
write_table_3dp <- function(df, out_path, sheet = "Sheet1") {
  wb <- createWorkbook()
  addWorksheet(wb, sheet)
  writeData(wb, sheet, df)
  
  numStyle3 <- createStyle(numFmt = "0.000")
  num_cols <- which(vapply(df, is.numeric, logical(1)))
  if (length(num_cols) > 0) {
    addStyle(
      wb, sheet, style = numStyle3,
      rows = 2:(nrow(df) + 1),
      cols = num_cols,
      gridExpand = TRUE, stack = TRUE
    )
  }
  
  setColWidths(wb, sheet, cols = 1:ncol(df), widths = "auto")
  saveWorkbook(wb, out_path, overwrite = TRUE)
}

# ---------- NORMALISE (qcrlscR LOESS) ----------
dat_corr <- qc.rlsc.wrap(
  dat    = data_o,
  cls.qc = cls_qc,
  cls.bl = cls_bl,
  method = "divide",
  intra  = TRUE,
  opti   = TRUE,
  log10  = TRUE,
  outl   = TRUE,
  shift  = FALSE
)
rownames(dat_corr) <- rownames(data_o)

# ---------- OUTPUTS ----------
cv_tbl <- data.frame(
  Feature            = feat_names,
  CV_all_Raw         = NA_real_,
  CV_pool_Raw        = NA_real_,
  CV_all_Corrected   = NA_real_,
  CV_pool_Corrected  = NA_real_,
  Diff_CV_all        = NA_real_,
  Diff_CV_pool       = NA_real_,
  WithinCV_Raw       = NA_real_,
  WithinCV_Corrected = NA_real_,
  Diff_WithinCV      = NA_real_,
  Improved_both      = FALSE,
  check.names = FALSE
)

pdf_out <- file.path(Data_normalisation_dir, "qcrlscR_LOESS_Normalisation.pdf")
pdf(pdf_out, width = 14, height = 5)

# ---------- LOOP FEATURES (CALC + PLOTS HERE) ----------
for (feature in feat_names) {
  
  raw_vec  <- data_o[[feature]];  names(raw_vec)  <- rownames(data_o)
  corr_vec <- dat_corr[[feature]]; names(corr_vec) <- rownames(dat_corr)
  
  # CVs: non-QC and pool
  cv_raw_all   <- cv_safe(raw_vec[samp_ids])
  cv_corr_all  <- cv_safe(corr_vec[samp_ids])
  cv_raw_pool  <- cv_safe(raw_vec[qc_samples])
  cv_corr_pool <- cv_safe(corr_vec[qc_samples])
  
  # within-group CVs on NON-QC samples
  wg_raw  <- within_cv_safe(raw_vec[samp_ids],  group_labels_samp)
  wg_corr <- within_cv_safe(corr_vec[samp_ids], group_labels_samp)
  
  # table
  cv_tbl[cv_tbl$Feature == feature,
         c("CV_all_Raw","CV_pool_Raw","CV_all_Corrected","CV_pool_Corrected")] <-
    c(cv_raw_all, cv_raw_pool, cv_corr_all, cv_corr_pool)
  
  cv_tbl[cv_tbl$Feature == feature, c("Diff_CV_all","Diff_CV_pool")] <-
    c(cv_corr_all - cv_raw_all, cv_corr_pool - cv_raw_pool)
  
  cv_tbl[cv_tbl$Feature == feature,
         c("WithinCV_Raw","WithinCV_Corrected","Diff_WithinCV")] <-
    c(wg_raw, wg_corr, wg_corr - wg_raw)
  
  cv_tbl[cv_tbl$Feature == feature, "Improved_both"] <-
    isTRUE(is.finite(cv_raw_all) && is.finite(cv_corr_all) && is.finite(wg_raw) && is.finite(wg_corr) &&
             cv_corr_all < cv_raw_all && wg_corr < wg_raw)
  
  # ---- FIT LINE SHAPE (consistent with QC-RLSC) ----
  tmp_raw <- data_o[, feature, drop = FALSE]
  colnames(tmp_raw) <- "feat"
  
  corr1 <- qc.rlsc(
    x = tmp_raw,
    y = cls_qc,
    method = "divide",
    opti = TRUE
  )
  
  fit_shape <- as.numeric(tmp_raw[, 1] / corr1[, 1])
  fit_shape[!is.finite(fit_shape)] <- NA_real_
  
  # scale to raw intensity scale
  k <- median(tmp_raw[qc_idx, 1], na.rm = TRUE) / median(fit_shape[qc_idx], na.rm = TRUE)
  fit_all_raw  <- fit_shape * k
  fit_all_corr <- rep(median(corr_vec[qc_idx], na.rm = TRUE), nrow(SSF_o))
  
  df_plot <- tibble(
    Sample    = factor(rownames(SSF_o), levels = rownames(SSF_o)),
    Group     = factor(SSF_o$Conditions, levels = unique(SSF_o$Conditions)),
    Raw       = raw_vec,
    Corrected = corr_vec,
    FitRaw    = fit_all_raw,
    FitCorr   = fit_all_corr
  )
  
  # subtitles: left=RAW, right=CORRECTED (include pool CV)
  sub_raw  <- sprintf("Raw: CV(sample)=%s%% | WithinCV=%s%% | CV(pool)=%s%%",
                      fmt1(cv_raw_all), fmt1(wg_raw), fmt1(cv_raw_pool))
  sub_corr <- sprintf("Corrected: CV(sample)=%s%% | WithinCV=%s%% | CV(pool)=%s%%",
                      fmt1(cv_corr_all), fmt1(wg_corr), fmt1(cv_corr_pool))
  
  p_raw <- ggplot(df_plot, aes(x = Sample, y = Raw, color = Group)) +
    geom_point(size = 2) +
    geom_line(aes(y = FitRaw, group = 1), color = "red", linewidth = 0.8) +
    scale_color_manual(values = customColors) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = paste0(feature, " — Raw"), subtitle = sub_raw) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  p_corr <- ggplot(df_plot, aes(x = Sample, y = Corrected, color = Group)) +
    geom_point(size = 2) +
    geom_line(aes(y = FitCorr, group = 1), color = "red", linewidth = 0.8) +
    scale_color_manual(values = customColors) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = paste0(feature, " — Corrected"), subtitle = sub_corr) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  gridExtra::grid.arrange(p_raw, p_corr, ncol = 2)
}

dev.off()

# ---------- EXPORT EXCEL (per-feature table) ----------
cv_outfile <- file.path(Data_normalisation_dir, "qcrlscR_LOESS_CV_summary.xlsx")
write_table_3dp(cv_tbl, cv_outfile)

# ---------- SUMMARY TABLE (like your three_col_tbl) ----------
# mean differences vs raw (more negative = better)
cv_mean_diff_all    <- mean(cv_tbl$Diff_CV_all,    na.rm = TRUE)
cv_mean_diff_within <- mean(cv_tbl$Diff_WithinCV,  na.rm = TRUE)
cv_mean_diff_pool   <- mean(cv_tbl$Diff_CV_pool,   na.rm = TRUE)

three_col_tbl <- data.frame(
  Method         = "qcrlscR_LOESS",
  CV             = round(cv_mean_diff_all, 4),
  Within_CV      = round(cv_mean_diff_within, 4),
  Pool_CV        = round(cv_mean_diff_pool, 4),
  Improved       = isTRUE(cv_mean_diff_all < 0 & cv_mean_diff_within < 0 & cv_mean_diff_pool < 0),
  check.names = FALSE
)

message("Mean difference vs Raw (more negative = better):")
print(three_col_tbl)

summary_outfile <- file.path(Data_normalisation_dir, "qcrlscR_LOESS_MeanDiff_fromRaw_CVtable.xlsx")
write_table_3dp(three_col_tbl, summary_outfile)

# ---------- corrected matrix (optional) ----------
corr_outfile <- file.path(Data_normalisation_dir, "qcrlscR_LOESS_corrected_matrix_3dp.xlsx")
write_table_3dp(
  df = as.data.frame(round(dat_corr, 3)) %>% tibble::rownames_to_column("Sample"),
  out_path = corr_outfile
)

message("Saved PDF: ", pdf_out)
message("Saved CV table: ", cv_outfile)
message("Saved summary table: ", summary_outfile)
message("Saved corrected matrix (3dp): ", corr_outfile)

# bring back order
data_list[[1]] <- dat_corr[rownames(SSF),]
```

    
#2.3. Missing value imputation
```{r} 
#### Make preMVI boxplots ---------------
Metab_AllValuePLots <- list()
for (metab in colnames(rbind(data_list[[1]], data_list[[2]]))) {   # metab = colnames(data)[1]
  # Merge metabolite values with metadata
  plotdata <- merge(rbind(data_list[[1]], data_list[[2]]) %>% select(all_of(metab)),  rbind(SSF_list[[1]], SSF_list[[2]]), by = 0)
  
  # Ensure RunOrder is numeric and sort
 # plotdata <- plotdata[order(plotdata$RunOrder), ]
  
  # Make Row.names a factor ordered by RunOrder
  plotdata$Row.names <- factor(plotdata$Row.names, levels = plotdata$Row.names)
  
  # Plot with Colors column used for custom colors but Conditions for legend
  p <- ggplot(plotdata, aes(x = Row.names, y = .data[[metab]], fill = Conditions)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = setNames(plotdata$Color, plotdata$Conditions)) + # map colors
    labs(title = paste(metab),
         x = "Samples (Run Order)", y = "Intensity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),  # rotate x labels
          plot.title = element_text(size = 14))
  
  # Store plot
  Metab_AllValuePLots[[metab]] <- p
}

# Save blank Bar-plots
AllValueBoxplotPreMVIloc <- file.path(output.dir,paste0("Metabolite_AllValues_preMVI_boxplots.pdf"))
pdf(AllValueBoxplotPreMVIloc, width = 14, height = 8) # Adjust width and height as needed

# Iterate over the list of plots and print each one
for (plot in Metab_AllValuePLots) {
  print(plot)
}

# Close the PDF device
dev.off()
    

### ----------------- Missign Value Imputation ------------------
for(i in 1:length(SSF_list)){ #i=1
  # Dataset selection
  SSF = SSF_list[[i]]
  data = data_list[[i]]
  if(i==1){
    outname<- "sample_"
  }else{
    outname<- "ctrl_"
  }
  
  MVI_data <- data 
  SSF <- SSF[match(rownames(MVI_data), rownames(SSF)), ]
  halfmin <- min(MVI_data[MVI_data>0], na.rm = T)/2
  # Because of the loess correction we have zeros and negative values along with the NAs we set them all to zero.
  # MVI_data[MVI_data <halfmin] <- 0
  # MVI_data[is.na(MVI_data)] <- 0
  
  for (feature  in colnames(MVI_data)){ # feature =  colnames(MVI_data)[3]
    feature_data <- merge(MVI_data[feature] , SSF %>% select("Conditions"), by= 0, sort = FALSE)
    feature_data <- column_to_rownames(feature_data, "Row.names")
    feature_data[feature_data==0] <- NA
    
    split_Input <- split(feature_data, SSF[,"Conditions"]) # split data frame into a list of data frames by condition
    
    for (m in split_Input){ # Select dataframe  # m <- split_Input[[2]]
      if( sum(is.na(m[,1]))== (length(m[,1])) |  sum(is.na(m[,1]))== (length(m[,1])-1) ){ #  if  all is NA or we have only 1 value then put 0 in whole group 
        set.seed(1234)
        for (value in  1:length( MVI_data[rownames(MVI_data) %in% rownames(m),feature])){ # value = 1
          MVI_data[rownames(MVI_data) %in% rownames(m),feature][value] <- halfmin + sample(100:1000, 1, replace = TRUE)
        }
      }else{ # if we have more than 1 value
        MVI_data[rownames(MVI_data) %in% rownames(m),feature][is.na(m[,1])]<- mean(m[,1],na.rm = TRUE )
      }
    }
  }   
  
  # set.seed(123) # Set seed for reproducibility
  # MVI_data[MVI_data == 0] <- 1 + runif(sum(MVI_data == 0), min = 0.01, max = 0.1)
  
  # PCA
  make_pca_grid(MVI_data, SSF, add_to_name =  paste0("7. ",outname,"Imputed"))
  make_signal_drift_reports(MVI_data, SSF, add_to_name =  paste0("7. ",outname,"Imputed"))
  
  data_list[[i]] <- MVI_data
}



###  Make postMVI boxplots -------------------
Metab_AllValuePLots <- list()
for (metab in colnames(rbind(data_list[[1]], data_list[[2]]))) {   # metab = colnames(data)[1]
  # Merge metabolite values with metadata
  plotdata <- merge(rbind(data_list[[1]], data_list[[2]]) %>% select(all_of(metab)),  rbind(SSF_list[[1]], SSF_list[[2]]), by = 0)
  
  # Ensure RunOrder is numeric and sort
 # plotdata <- plotdata[order(plotdata$RunOrder), ]
  
  # Make Row.names a factor ordered by RunOrder
  plotdata$Row.names <- factor(plotdata$Row.names, levels = plotdata$Row.names)
  
  # Plot with Colors column used for custom colors but Conditions for legend
  p <- ggplot(plotdata, aes(x = Row.names, y = .data[[metab]], fill = Conditions)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = setNames(plotdata$Color, plotdata$Conditions)) + # map colors
    labs(title = paste(metab),
         x = "Samples (Run Order)", y = "Intensity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),  # rotate x labels
          plot.title = element_text(size = 14))
  
  # Store plot
  Metab_AllValuePLots[[metab]] <- p
}

# Save blank Bar-plots
AllValueBoxplotPreMVIloc <- file.path(output.dir,paste0("Metabolite_AllValues_postMVI_boxplots.pdf"))
pdf(AllValueBoxplotPreMVIloc, width = 14, height = 8) # Adjust width and height as needed

# Iterate over the list of plots and print each one
for (plot in Metab_AllValuePLots) {
  print(plot)
}

# Close the PDF device
dev.off()

paste("#2.3. Missing value imputation Finished!")
```


# CoRe normalisation using media blanks
```{r}
# -------------------- SETUP --------------------
norm_dir <- file.path(Data_normalisation_dir)
if (!dir.exists(norm_dir)) dir.create(norm_dir, recursive = TRUE)

if (CoRe == TRUE) {
  

  # -------------------- INPUT --------------------
  media_blank_data <- data_list[[2]]
  media_blank_ssf  <- SSF_list[[2]]
  
  sample_data <- data_list[[1]]
  sample_ssf  <- SSF_list[[1]]
  
  # Keep SSF aligned to sample_data rows
  common_samples <- intersect(rownames(sample_data), rownames(sample_ssf))
  sample_data <- sample_data[common_samples, , drop = FALSE]
  sample_ssf  <- sample_ssf[common_samples, , drop = FALSE]
  
  # -------------------- BLANK MEANS (per blank condition) --------------------
  blank_levels <- unique(media_blank_ssf$Conditions)
  
  blank_means <- sapply(blank_levels, function(g) {
    colMeans(media_blank_data[media_blank_ssf$Conditions == g, , drop = FALSE], na.rm = TRUE)
  })
  
  blank_means_df <- as.data.frame(t(blank_means))
  rownames(blank_means_df) <- blank_levels
  
  # IMPORTANT: do NOT convert NA -> 0 here (hides missing blank info)
  
  # -------------------- MATCHING: sample condition -> blank condition --------------------
  pick_blank_group <- function(sample_cond, blank_levels) {
    sample_cond  <- as.character(sample_cond)
    blank_levels <- as.character(blank_levels)
    
    if (is.na(sample_cond) || !nzchar(sample_cond)) return(NA_character_)
    
    s  <- tolower(sample_cond)
    bl <- tolower(blank_levels)
    
    if (s %in% bl) return(blank_levels[match(s, bl)])
    
    hits <- which(vapply(bl, function(b) grepl(b, s, fixed = TRUE) || grepl(s, b, fixed = TRUE), logical(1)))
    
    if (length(hits) == 1) return(blank_levels[hits])
    
    if (length(hits) > 1) {
      lens <- nchar(bl[hits])
      return(blank_levels[hits[which.max(lens)]])
    }
    
    NA_character_
  }
  
  # -------------------- SUBTRACT BLANK MEANS --------------------
  sample_data_corrected <- as.data.frame(
    matrix(NA_real_, nrow = nrow(sample_data), ncol = ncol(sample_data),
           dimnames = list(rownames(sample_data), colnames(sample_data)))
  )
  
  for (i in seq_len(nrow(sample_data))) {
    
    cond <- sample_ssf$Conditions[i]
    
    # If pool: keep as NA (or keep original), but your later code removes pools anyway
    if (!is.na(cond) && tolower(cond) == "pool") {
      sample_data_corrected[i, ] <- NA_real_
      next
    }
    
    bg <- pick_blank_group(cond, rownames(blank_means_df))
    
    if (is.na(bg)) {
      warning(sprintf("No blank group match for sample condition: %s (row %d). Setting row to NA.", cond, i))
      sample_data_corrected[i, ] <- NA_real_
      next
    }
    
    # subtract; keeps NA where sample has NA
    sample_data_corrected[i, ] <- sample_data[i, ] - blank_means_df[bg, ]
  }
  
  # -------------------- AMOUNT NORMALISATION (multiply by 1/amount) --------------------
  SSF <- sample_ssf
  
  # SSF$Amount_numeric <- suppressWarnings(as.numeric(SSF$`Amount of Material`))
  # amounts <- SSF$Amount_numeric
  # names(amounts) <- rownames(SSF)
  # amounts[is.na(amounts) | amounts == 0] <- 1
  # amounts <- 1 / amounts
  
  # align amounts to corrected matrix rows
  # amounts <- amounts[rownames(sample_data_corrected)]
  # amounts[is.na(amounts)] <- 1
  # 
  # # multiply each row by amounts (your intended behaviour)
  # sample_data_corrected <- as.data.frame(sweep(sample_data_corrected, 1, amounts, FUN = "*"))
  
    # align growth table to data by rownames
  gf <- ssf_for_growth_factor[rownames(sample_data_corrected), "Cell Count at Experimental End Point"]

  # n_cells_at_start <- 500000
  # 
  # gf <- gf/n_cells_at_start 
  
  gf[is.na(gf)] <- 1

  
  # divide each sample (row) by its growth factor
  sample_data_corrected <- sweep(sample_data_corrected, 1, gf, FUN = "/")


  
  # -------------------- QC PLOTS (RAW vs BLANK MEAN BARS) --------------------
  pdf(file = file.path(norm_dir, "Media_data_with_CtrlMediaBar2.pdf"), width = 12, height = 5)
  
  for (metab in colnames(sample_data_corrected)) {
    
    plot_df <- tibble(
      Sample = rownames(sample_data_corrected),
      Value  = sample_data[[metab]],          # keeping your original choice (raw shown as points)
      Group  = SSF$Conditions
    )
    
    blank_df <- tibble(
      Group_media = rownames(blank_means_df),
      MeanValue   = blank_means_df[[metab]]
    ) %>%
      mutate(Group = gsub("_control$", "_sample", Group_media))
    
    p1 <- ggplot() +
      geom_col(
        data = blank_df,
        aes(x = Group, y = MeanValue),
        fill = "grey", alpha = 0.6
      ) +
      geom_jitter(
        data = plot_df,
        aes(x = Group, y = Value),
        width = 0.15, size = 2, colour = "blue", alpha = 0.7
      ) +
      theme_bw() +
      labs(title = metab, x = "Condition", y = "Intensity") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
    print(p1)
  }
  
  dev.off()
  
  # -------------------- QC PLOTS (RAW vs BLANK MEAN BARS) - NO POOLS --------------------
pdf(file = file.path(norm_dir, "Media_data_with_CtrlMediaBar2_noPools.pdf"),
    width = 12, height = 5)

for (metab in colnames(sample_data_corrected)) {

  plot_df <- tibble::tibble(
    Sample = rownames(sample_data_corrected),
    Value  = sample_data[[metab]],          # raw points (as in your original)
    Group  = SSF$Conditions
  ) %>%
    dplyr::filter(!is.na(Group)) %>%
    dplyr::filter(!grepl("pool", Group, ignore.case = TRUE))  # <-- remove pools

  blank_df <- tibble::tibble(
    Group_media = rownames(blank_means_df),
    MeanValue   = blank_means_df[[metab]]
  ) %>%
    dplyr::mutate(Group = gsub("_control$", "_sample", Group_media))

  # keep only x categories that actually appear (either as sample points or blank bars)
  x_levels <- union(unique(plot_df$Group), unique(blank_df$Group))
  plot_df$Group  <- factor(plot_df$Group, levels = x_levels)
  blank_df$Group <- factor(blank_df$Group, levels = x_levels)

  p1 <- ggplot2::ggplot() +
    ggplot2::geom_col(
      data = blank_df,
      ggplot2::aes(x = Group, y = MeanValue),
      fill = "grey", alpha = 0.6
    ) +
    ggplot2::geom_jitter(
      data = plot_df,
      ggplot2::aes(x = Group, y = Value),
      width = 0.15, size = 2, colour = "blue", alpha = 0.7
    ) +
    ggplot2::theme_bw() +
    ggplot2::labs(title = metab, x = "Condition", y = "Intensity") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

  print(p1)
}

dev.off()
  
  # ---------- CoRe plotting: corrected data only (no grey bars), zero dotted line ----------
  pdf(file = file.path(norm_dir, "CoRe_CorrectedData_only.pdf"), width = 12, height = 5)
  
  # map sample -> condition (safe even if orders differ)
  cond_map <- SSF$Conditions
  names(cond_map) <- rownames(SSF)
  
  for (metab in colnames(sample_data_corrected)) {
    
    plot_df <- tibble::tibble(
      Sample = rownames(sample_data_corrected),
      Value  = as.numeric(sample_data_corrected[[metab]])
    ) %>%
      dplyr::mutate(Group = unname(cond_map[Sample])) %>%   # <- key fix
      dplyr::filter(!is.na(Group)) %>%
      dplyr::filter(!grepl("pool", Group, ignore.case = TRUE))
    
    p1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = Group, y = Value)) +
      ggplot2::geom_jitter(width = 0.15, size = 2, colour = "blue", alpha = 0.7) +
      ggplot2::geom_hline(yintercept = 0, linetype = "dotted", colour = "black") +
      ggplot2::theme_bw() +
      ggplot2::labs(title = metab, x = "Condition", y = "Corrected intensity") +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
    
    print(p1)
  }
  
  dev.off()
  
  # -------------------- OUTPUT OBJECTS --------------------
  
  # remove pools (as in your code)
  SSF <- SSF[-grep("Pool", rownames(SSF), ignore.case = TRUE), , drop = FALSE]
  sample_data_corrected <- sample_data_corrected[-grep("Pool", rownames(sample_data_corrected), ignore.case = TRUE), , drop = FALSE]
  
  make_pca_grid(sample_data_corrected, SSF, add_to_name = "7.5 CoRe_BackgroundSubtracted")
  make_signal_drift_reports(sample_data_corrected, sample_ssf, add_to_name = "7.5 CoRe_BackgroundSubtracted")
}

###  Make postCoRe norm boxplots -------------------
Metab_AllValuePLots <- list()
for (metab in colnames(sample_data_corrected)) {   # metab = colnames(data)[1]
  # Merge metabolite values with metadata
  plotdata <- merge(sample_data_corrected %>% select(all_of(metab)),  SSF, by = 0)
  
  # Ensure RunOrder is numeric and sort
 # plotdata <- plotdata[order(plotdata$RunOrder), ]
  
  # Make Row.names a factor ordered by RunOrder
  plotdata$Row.names <- factor(plotdata$Row.names, levels = plotdata$Row.names)
  
  # Plot with Colors column used for custom colors but Conditions for legend
  p <- ggplot(plotdata, aes(x = Row.names, y = .data[[metab]], fill = Conditions)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = setNames(plotdata$Color, plotdata$Conditions)) + # map colors
    labs(title = paste(metab),
         x = "Samples (Run Order)", y = "Intensity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),  # rotate x labels
          plot.title = element_text(size = 14))
  
  # Store plot
  Metab_AllValuePLots[[metab]] <- p
}

# Save blank Bar-plots
AllValueBoxplotPreMVIloc <- file.path(output.dir,paste0("Metabolite_AllValues_postCoRenorm_boxplots.pdf"))
pdf(AllValueBoxplotPreMVIloc, width = 14, height = 8) # Adjust width and height as needed

# Iterate over the list of plots and print each one
for (plot in Metab_AllValuePLots) {
  print(plot)
}

# Close the PDF device
dev.off()

MVI_data <- sample_data_corrected 
paste("# CoRe normalisation using media blanks Finished!")
```


### --- 3. Data preparation/finalisation
#3.0 Preparation
```{r}
# Set sub folders
Data_finalisation_dir <- file.path(output.dir, "3. Data Finalisation")
if (!dir.exists(Data_finalisation_dir)) dir.create(Data_finalisation_dir, recursive = TRUE)

# Metabolite_Metadata$`SNratio<3` <- rownames(Metabolite_Metadata) %in% gsub("[^A-Za-z0-9 ]", "", names(blank_dominated_features[blank_dominated_features]))
# Data_cleaning_dir
# , ProjectCode,"_",Sys.Date(), 
paste("#3.0 Preparation Finished!")
```

#3.1 Data transformation
```{r}
if(CoRe==F){
  data <- log2(MVI_data) 
}else{
  data <-MVI_data
}
# PCA
PCA_data <- data
PCA_SSF <- SSF
make_pca_grid(PCA_data, PCA_SSF, add_to_name = "8. transformed")
make_signal_drift_reports(PCA_data, PCA_SSF, add_to_name = "8. transformed")
  
paste("#3.1 Data transformation Finished!")
```

#3.2. Outlier samples
```{r}
SSFsave <- SSF
# SSF <- SSFsave
initial_data <- data
data_norm <- data
Outlier_filtering_loop <- 2

sample_outliers <- list()
outlier_plot_list <- list()

for (loop in 1:Outlier_filtering_loop) {
  PCA.res <- prcomp(data_norm, center = TRUE, scale. = TRUE)
  
  # Scree plot data
  inflect_df <- data.frame(x = 1:length(PCA.res$sdev))
  inflect_df$y <- summary(PCA.res)$importance[2, ]
  inflect_df$Cumulative <- summary(PCA.res)$importance[3, ]
  screeplot_cumul <- format(round(inflect_df$Cumulative[1:20]*100, 1), nsmall = 1)
  knee <- inflection::uik(inflect_df$x, inflect_df$y)
  npcs <- knee - 1
  if(npcs <= 1) { npcs <- 2 }
  
  # Scree plot
  screeplot <- factoextra::fviz_screeplot(PCA.res, main = paste("PCA Explained variance plot filtering round ", loop),
                                          addlabels = TRUE, ncp = 20, geom = c("bar", "line"),
                                          barfill = "grey", barcolor = "grey", linecolor = "black") +
    theme_classic() +
    geom_vline(xintercept = npcs + 0.5, linetype = 2, color = "red") +
    annotate("text", x = 1:20, y = -0.8, label = screeplot_cumul, col = "black", size = 3)
  
  
  outlier_plot_list[[paste0("Scree_", loop)]] <-screeplot
  
  
  # PCA plot PC1 vs PC2
  pca_df <- data.frame(Sample = rownames(PCA.res$x), PCA.res$x[, 1:2], Group = SSF$Conditions)
  colnames(pca_df)[2:3] <- c("PC1", "PC2")
  
  pc1_var <- round(100 * summary(PCA.res)$importance[2, 1], 1)
  pc2_var <- round(100 * summary(PCA.res)$importance[2, 2], 1)
  
  pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
    geom_point(size = 3) +
    geom_text_repel(aes(label = Sample), size = 3, max.overlaps = 100) +
    scale_color_manual(values = setNames(SSF$Color, SSF$Conditions)) +
    labs(title = paste("PCA Score Plot Round", loop),
         x = paste0("PC1 (", pc1_var, "%)"),
         y = paste0("PC2 (", pc2_var, "%)")) +
    theme_classic()
  
  outlier_plot_list[[paste0("PCA_", loop)]] <-pca_plot
  
  # Hotelling's T2 test
  data_hot <- PCA.res$x[, 1:npcs]
  qcc_res <- qcc::mqcc(data_hot, type = "T2.single", labels = rownames(data_hot), confidence.level = 0.99, plot = FALSE)
  limits <- as.data.frame(qcc_res$limits)
  t2_data <- data.frame(Samples = names(qcc_res$statistics), Statistic = qcc_res$statistics)
  outliers <- t2_data[t2_data$Statistic > limits$UCL, , drop = FALSE]
  
  # Hotelling plot
  hotelling_plot <- ggplot(t2_data, aes(x = Samples, y = Statistic)) +
    geom_point(color = "blue", size = 2) +
    geom_point(data = outliers, aes(x = Samples, y = Statistic), color = "red", size = 3) +
    geom_line(group = 1, linetype = 2) +
    geom_hline(yintercept =limits[1,2], color = "red", linetype = "dashed") +
    labs(title = paste("Hotelling T2 Test Round", loop), y = "Group Summary Statistic") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  outlier_plot_list[[paste0("Hotelling_", loop)]] <- hotelling_plot
  
  # Remove outliers
  if (nrow(outliers) == 0) break
  data_norm <- data_norm[!rownames(data_norm) %in% outliers$Samples, , drop = FALSE]
  SSF <- SSF[!rownames(SSF) %in% outliers$Samples,]
  sample_outliers[[paste0("Round_", loop)]] <- outliers$Samples
}


Outlierloc <- file.path(Data_finalisation_dir,paste0("Outlier_testing" ,".pdf"))
pdf(Outlierloc, width = 8, height = 8) # Adjust width and height as needed

# Iterate over the list of plots and print each one
for (plot in outlier_plot_list) {
  print(plot)
}
# Close the PDF device
dev.off()

# Export outliers
if (length(sample_outliers) > 0) {
  sink(file.path(Data_finalisation_dir, "Outliers.txt"))
  print(sample_outliers)
  sink()
}

# data_norm <- sample_data_corrected

# # remove manually
# remove <- "ACNe01_Cells_017"
# data_norm <- data_norm[!rownames(data_norm) %in% remove, , drop = FALSE]
# SSF <- SSF[!rownames(SSF) %in% remove,]

# PCA
PCA_data <- data_norm
PCA_SSF <- SSF
res <- make_pca_grid(PCA_data, PCA_SSF, add_to_name = "9. Outlier_removed")
make_signal_drift_reports(PCA_data, PCA_SSF, add_to_name = "9. Outlier_removed")

paste("#3.2. Outlier samples Finished!")
```

# Merge technical replicates
```{r eval=FALSE}

data <- data_norm   # or MVI_data
SSF2 <- SSF

# align
common <- intersect(rownames(data), rownames(SSF2))
data <- data[common, , drop = FALSE]
SSF2 <- SSF2[common, , drop = FALSE]

# keep factor order
SSF2$Conditions <- factor(SSF2$Conditions, levels = levels(SSF$Conditions))

# -------------------- MERGE TECH REPS (median) --------------------
SSF2_tbl <- SSF2 %>% rownames_to_column("SampleID")

merged_tbl <- SSF2_tbl %>%
  group_by(Conditions, `Biological Replicate`) %>%
  summarise(
    merged = list(apply(data[SampleID, , drop = FALSE], 2, median, na.rm = TRUE)),
    Color  = dplyr::first(Color),
    .groups = "drop"
  ) %>%
  mutate(NewSampleID = paste0(Conditions, "_BioRep", `Biological Replicate`))

merged_data <- do.call(rbind, merged_tbl$merged)
rownames(merged_data) <- merged_tbl$NewSampleID
merged_data <- as.data.frame(merged_data)

merged_SSF <- merged_tbl %>%
  select(NewSampleID, Conditions, `Biological Replicate`, Color) %>%
  column_to_rownames("NewSampleID")

merged_SSF$Conditions

# -------------------- QC PLOTS (before vs merged) --------------------
pdf(file.path(output.dir, "Merge_TechReps_QCplots.pdf"), width = 10, height = 6)

for (metab in colnames(data)) {
  
  df_raw <- SSF2_tbl %>%
    mutate(
      Value = data[SampleID, metab],
      Type = "Raw"
    ) %>%
    select(SampleID, Conditions, `Biological Replicate`, Value, Type)
  
  df_merged <- merged_SSF %>%
    rownames_to_column("SampleID") %>%
    mutate(
      Value = merged_data[SampleID, metab],
      Type = "Merged"
    ) %>%
    select(SampleID, Conditions, `Biological Replicate`, Value, Type)
  
  df_plot <- bind_rows(df_raw, df_merged) %>%
    mutate(Type = factor(Type, levels = c("Raw", "Merged")))
  
  p <- ggplot(df_plot, aes(x = Type, y = Value, fill = Type)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.15, size = 1.8, alpha = 0.7) +
    facet_grid(`Biological Replicate` ~ Conditions, scales = "free_y") +
    theme_bw() +
    labs(title = metab, x = "", y = "Intensity") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
}

dev.off()

paste("#3.3 Technical Replicate Merging Finished!")
```


#3.3 Finalise data processing
```{r}
suppressMessages(in_data <- readxl::read_excel(file_listTF[grep("ssf", tolower(file_listTF), invert = T)], sheet = 1, col_names = T) |> as.data.frame()) # load data file
rownames(in_data) <- in_data[[1]]   # set first column as row names
in_data <- in_data[, -1]     
write.csv(in_data, paste0(Data_finalisation_dir, "/Raw_Data.csv"))


processed_data <- data_norm
#processed_data$`valine-d8` <- NULL
processed_data <- processed_data[,-grep(valine_pattern_id, tolower(colnames(processed_data)))]

processed_data <- processed_data[!grepl("pool", tolower(rownames(processed_data))),]
processed_data[sapply(processed_data, is.numeric)] <- round(processed_data[sapply(processed_data, is.numeric)], 3)

out_data <- processed_data
SSF <- SSF[rownames(out_data),]

write.csv(out_data, paste0(Data_finalisation_dir, "/Processed_Data.csv"))


if (exists("merge_tectno_Reps") && isTRUE(merge_tectno_Reps)) {
  SSF <- merged_SSF
  
  processed_data <- data_norm
#processed_data$`valine-d8` <- NULL
processed_data <- processed_data[,-grep(valine_pattern_id, tolower(colnames(processed_data)))]

processed_data <- processed_data[!grepl("pool", tolower(rownames(processed_data))),]
processed_data[sapply(processed_data, is.numeric)] <- round(processed_data[sapply(processed_data, is.numeric)], 3)

out_data <- processed_data
SSF <- SSF[rownames(out_data),]

write.csv(out_data, paste0(Data_finalisation_dir, "/Processed_Data_unmerged.csv"))

processed_data <- merged_data
#processed_data$`valine-d8` <- NULL
processed_data <- processed_data[,-grep(valine_pattern_id, tolower(colnames(processed_data)))]

processed_data <- processed_data[!grepl("pool", tolower(rownames(processed_data))),]
processed_data[sapply(processed_data, is.numeric)] <- round(processed_data[sapply(processed_data, is.numeric)], 3)

out_data <- processed_data
merged_SSF <- merged_SSF[rownames(out_data),]

write.csv(out_data, paste0(Data_finalisation_dir, "/Processed_Data.csv"))
}

# PCA
PCA_data <- processed_data
PCA_SSF <- SSF
res <- make_pca_grid(PCA_data, PCA_SSF, add_to_name = "10. Processed_data")

#make_signal_drift_reports(PCA_data, PCA_SSF, add_to_name = "10. Processed_data")

paste("#3.3 Finalise data processing Finished!")
```


# CHeck if BIOlogical vairation is in first 2 compoents
```{r eval = FALSE}

# -------------------- INPUT --------------------
X <- as.matrix(processed_data)          # samples in rows, features in columns
meta <- SSF                     # sample metadata; rownames must match MVI_data

# keep only common samples + align
common <- intersect(rownames(X), rownames(meta))
X <- X[common, , drop = FALSE]
meta <- meta[common, , drop = FALSE]

# optional: log transform (skip if already log)
# X <- log10(X + 1)

# remove rows with any NA / Inf
keep <- is.finite(rowSums(X)) & !apply(X, 1, function(r) any(!is.finite(r)) || any(is.na(r)))
X <- X[keep, , drop = FALSE]
meta <- meta[keep, , drop = FALSE]

# -------------------- PCA --------------------
pca <- prcomp(X, center = TRUE, scale. = TRUE)

scores <- as.data.frame(pca$x[, 1:3])
scores$Sample <- rownames(scores)

df <- scores %>%
  left_join(meta %>% rownames_to_column("Sample"), by = "Sample")

# % variance for axis labels
pc_var <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 2)

# -------------------- PLOT --------------------
p <- ggplot(df, aes(PC1, PC2,
                    fill  = factor(`Biological Replicate`),
                    shape = Conditions)) +
  geom_point(size = 5, stroke = 1, color = "black") +
  scale_shape_manual(values = rep(c(21,22,23,24,25),
                                  length.out = length(unique(df$Conditions)))) +
  guides(
    fill = guide_legend(override.aes = list(shape = 21, color = "black", size = 5))
  ) +
  theme_classic() +
  labs(
    fill  = "Biological Replicate",
    shape = "Condition",
    x = paste0("PC1 (", pc_var[1], "%)"),
    y = paste0("PC2 (", pc_var[2], "%)")
  )

ggsave(file.path(output.dir, "PCA_BiologicalReplicate.svg"),
       plot = p, width = 12, height = 10)
```

### ----- 4. Data Visualisations ----- 
#4.1. Final PCA
```{r}
#update frpm original  colors
customColors <- customColorsBox

data <-processed_data

# --- 1. Define color column and ensure output folder exists ---
cols_to_color <- colnames(SSF)[1]

# (Optional) skip columns with too many unique values (e.g., IDs)
# cols_to_color <- cols_to_color[sapply(cols_to_color, function(c)
#   length(unique(SSF[[c]])) <= 30)]

# Run your function for each SSF column
for (cc in cols_to_color) {
  make_pca_grid(
    data        = data,
    SSF         = SSF,
    add_to_name = paste0("Final PCA_by_", cc),
    project_code= ProjectCode,
    color_col   = cc,
    fig_size_cm = c(60, 40)
  )
}


# --- 3. Perform PCA once on all samples (Pools included) ---
pca_res <- prcomp(data, scale. = TRUE)
pca_var_explained <- summary(pca_res)$importance[2, 1:5] * 100  # % for PC1–PC5
SSF$Label <- gsub(paste0(ProjectCode, "_"), "", rownames(SSF))  # Sample names for plotting

# --- 6. Final PCA plot with PC1 vs PC2 excluding Pools ---
non_pool <- !grepl("pool", tolower(SSF$Conditions))
pca_df <- data.frame(pca_res$x[non_pool, 1:2],
                     Conditions = SSF$Conditions[non_pool],
                     Label = SSF$Label[non_pool])
colnames(pca_df)[1:2] <- c("PC1", "PC2")

pc1_var <- round(100 * summary(pca_res)$importance[2, 1], 1)
pc2_var <- round(100 * summary(pca_res)$importance[2, 2], 1)

# make sure factor levels match palette order (optional but tidy)
cond_levels <- intersect(unique(pca_df$Conditions), customColors$Conditions)
pca_df$Conditions <- factor(pca_df$Conditions, levels = cond_levels)
pal_use <- customColors$Color[match(cond_levels, customColors$Conditions)]

final_pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Conditions)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.2) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.2) +
  scale_color_manual(values = pal_use) +
  labs(title = "PCA: PC1 vs PC2 by Condition", x = paste0("PC1 (", pc1_var, "%)"),
       y = paste0("PC2 (", pc2_var, "%)")) +
  theme_classic()


ggsave(file.path(file.path(output.dir, "PCA_plots"), "Final_PCA_by_Conditions.svg"),
       plot = final_pca_plot, width = 20, height = 15, units = "cm")


# --- Add names to final PCA ---

final_pca_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Conditions)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = Label), size = 3, show.legend = FALSE) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.2) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.2) +
  scale_color_manual(values = pal_use) +
  labs(
    title = "PCA: PC1 vs PC2 by Condition",
    x = paste0("PC1 (", pc1_var, "%)"),
    y = paste0("PC2 (", pc2_var, "%)")
  ) +
  theme_classic()

ggsave(file.path(file.path(output.dir, "PCA_plots"), "Final_PCA_by_Conditions_Labeled.svg"),
       plot = final_pca_plot, width = 20, height = 15, units = "cm")
paste("#4.1. Final PCA Finished!")
```

#4.2. Heatmap
```{r} 
# Custom scaling function: scale values to [-1, 1]
scale_to_minus_one_to_one <- function(vec) {
  if (max(vec, na.rm = TRUE) == min(vec, na.rm = TRUE)) {
    return(rep(0, length(vec)))  # If constant, return 0s
  }
  2 * (vec - min(vec, na.rm = TRUE)) / (max(vec, na.rm = TRUE) - min(vec, na.rm = TRUE)) - 1
}

# Scale data
scaled_data <- as.data.frame(lapply(processed_data, scale_to_minus_one_to_one))   # or out_data
colnames(scaled_data) <- colnames(processed_data)
rownames(scaled_data) <- rownames(processed_data)

# Remove pools
# scaled_data <- scaled_data[-grep("pool", tolower(rownames(scaled_data))),]

# --------------------- CHANGED ANNOTATION ONLY ---------------------

# Sample annotation (align to heatmap columns = rownames(scaled_data))
annotation_col <- data.frame(Condition = SSF$Conditions, row.names = rownames(SSF))
annotation_col <- annotation_col[rownames(scaled_data), , drop = FALSE]
annotation_col$Condition <- as.factor(annotation_col$Condition)

# Build color mapping from SSF$Color
pal_df   <- unique(SSF[, c("Conditions", "Color")])
cond_pal <- setNames(pal_df$Color, pal_df$Conditions)   # named vector: names = condition, values = color

annotation_colors <- list(Condition = cond_pal)

# ------------------------------------------------------------------

# Feature annotation with safe rownames
clean_for_match <- function(x) gsub("[*†]", "", x)

orig_names  <- colnames(scaled_data)           # keep symbols for display
clean_names <- clean_for_match(orig_names)     # remove symbols for matching

# Match to universe
annotation_row <- data.frame(MainClass = universe$`Main.class`)
rownames(annotation_row) <- universe$`Input.name`
annotation_row <- annotation_row[clean_names, , drop = FALSE]

# Assign original names with symbols as rownames (display)
rownames(annotation_row) <- orig_names

# Ensure annotation_row only contains rows matching the final matrix
annotation_row <- annotation_row[colnames(scaled_data), , drop = FALSE]

# Plot
svg(paste0(output.dir, "/Heatmap_scaled.svg"), width = 10, height = 22)
set.seed(1234)
pheatmap(
  mat = t(as.matrix(scaled_data)),
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  annotation_col = annotation_col,
  annotation_row = annotation_row,
  annotation_colors = annotation_colors,
  main = "Metabolite heatmap",
  angle_col = 45,
  fontsize_col = 8,
  fontsize_row = 6
)
dev.off()

# ---------------- HEATMAP 2 (samples grouped by SSF$Conditions levels) ----------------

# enforce group order = factor levels
annotation_col$Condition <- factor(annotation_col$Condition, levels = levels(SSF$Conditions))

# order samples by Condition, keeping within-group order as-is
ord <- order(annotation_col$Condition)
scaled_data2 <- scaled_data[ord, , drop = FALSE]
annotation_col2 <- annotation_col[ord, , drop = FALSE]

svg(paste0(output.dir, "/Heatmap_Unclustered_Condition.svg"), width = 10, height = 22)
set.seed(1234)
pheatmap(
  mat = t(as.matrix(scaled_data2)),
  cluster_rows = TRUE,     # <-- only change: remove row clustering too
  cluster_cols = FALSE,
  clustering_method = "ward.D2",
  clustering_distance_rows = "euclidean",
  # clustering_distance_cols = "euclidean",
  annotation_col = annotation_col2,
  annotation_row = annotation_row,
  annotation_colors = annotation_colors,
  main = "Metabolite heatmap Unclustered Conditions",
  angle_col = 45,
  fontsize_col = 8,
  fontsize_row = 6
)
dev.off()

paste("#4.2. Heatmap Finished!")

```

#4.3. Boxplots
```{r} 
Box_data <- processed_data #processed_data # the log2 data --> out_data

Box_data <- merge( SSF %>%select(Conditions),Box_data, by=0)
Box_data <- column_to_rownames(Box_data, "Row.names")
Box_data <- Box_data[!grepl("Pool", Box_data$Conditions),]
Box_data$Conditions <- factor(Box_data$Conditions,levels =   unique(unique(SSF$Conditions)  )) # Merge for boxplots we use the SSSF check it out its potential reason for error!!
boxplot_colors <- customColorsBox[match(levels(Box_data$Conditions), customColorsBox$Conditions), ]

plot_list <- list()
for (i in 2:ncol(Box_data)){ # i=4
  message( colnames(Box_data[i]))
  
  boxplotdata <- Box_data %>%  select(i,1) %>%                       
    group_by(Conditions)
  NAMES <- names(boxplotdata)[1]
  message("working metabolite: ",NAMES)
  names(boxplotdata)[1] <- c("Intensity")
  
  
  extract <- boxplotdata %>% as.data.frame()
  extract$Intensity <- as.numeric(extract$Intensity)
  
  #extract$Conditions <- factor(extract$Conditions, levels = names(boxplot_colors))
  box_base <- ggbetweenstats(data = extract,
                             x = Conditions,
                             y = Intensity,
                             bf.message = FALSE,
                             title = paste(NAMES),
                             results.subtitle = FALSE,
                             pairwise.comparisons = FALSE,
                             pairwise.display = "none",
                             ylab = "normalised intensity"
  )
  
  # Force y-axis to start at 0
  boxplot <- box_base +
    geom_hline(yintercept = 0, colour = "black", linetype = "dotted") +
    ggplot2::scale_color_manual(values = c(boxplot_colors$Color))+
    # Unsilence the line below to stop forcing zero on y
    #scale_y_continuous(limits = range(extract$Intensity, na.rm = TRUE))+
    theme(
      plot.title = element_text(size = 22),
      plot.subtitle = element_text(size = 10),
      legend.title = element_text(size = 20),
      legend.text = element_text(size = 20),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10)
    )

  

  
  plot_list[NAMES][[1]] <- boxplot #final_plot 
}


Boxplotloc <- file.path(output.dir,paste0("Boxplots_Y=0" ,".pdf"))
pdf(Boxplotloc, width = 8, height = 8) # Adjust width and height as needed

# Iterate over the list of plots and print each one
for (plot in plot_list) {
  print(plot)
}

# Close the PDF device
dev.off()



plot_list <- list()
for (i in 2:ncol(Box_data)){ # i=4
  message( colnames(Box_data[i]))
  
  boxplotdata <- Box_data %>%  select(i,1) %>%                       
    group_by(Conditions)
  NAMES <- names(boxplotdata)[1]
  message("working metabolite: ",NAMES)
  names(boxplotdata)[1] <- c("Intensity")
  
  
  extract <- boxplotdata %>% as.data.frame()
  extract$Intensity <- as.numeric(extract$Intensity)
  
  #extract$Conditions <- factor(extract$Conditions, levels = names(boxplot_colors))
  box_base <- ggbetweenstats(data = extract,
                             x = Conditions,
                             y = Intensity,
                             bf.message = FALSE,
                             title = paste(NAMES),
                             results.subtitle = FALSE,
                             pairwise.comparisons = FALSE,
                             pairwise.display = "none",
                             ylab = "normalised intensity"
  )
  
  # Force y-axis to start at 0
  boxplot <- box_base +
    geom_hline(yintercept = 0, colour = "black", linetype = "dotted") +
    ggplot2::scale_color_manual(values = c(boxplot_colors$Color))+
    # Unsilence the line below to stop forcing zero on y
    scale_y_continuous(limits = range(extract$Intensity, na.rm = TRUE))+
    theme(
      plot.title = element_text(size = 22),
      plot.subtitle = element_text(size = 10),
      legend.title = element_text(size = 20),
      legend.text = element_text(size = 20),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10)
    )

  

  
  plot_list[NAMES][[1]] <- boxplot #final_plot 
}


Boxplotloc <- file.path(output.dir,paste0("Boxplots" ,".pdf"))
pdf(Boxplotloc, width = 8, height = 8) # Adjust width and height as needed

# Iterate over the list of plots and print each one
for (plot in plot_list) {
  print(plot)
}

# Close the PDF device
dev.off()
paste("#4.3. Boxplots Finished!")
```



### ------ 5. Data Analysis ------
#5.1.DMA and Ora parameter setting
```{r}
DMA_dir <- file.path(output.dir, "4. DMA")
if (!dir.exists(DMA_dir)) dir.create(DMA_dir, recursive = TRUE)

unique(SSF$Conditions[match(rownames(processed_data), rownames(SSF))])
# 3. Get two groups
group1 <-"Type_2"   # Change as needed
group2 <- "Vehicle"   # Change as needed

comp_name <- paste0(group1, "_Vs_", group2)

subdir_path <- file.path(DMA_dir, comp_name)
# 2. Create the subdirectory if it doesn't exist
if (!dir.exists(subdir_path)) {
  dir.create(subdir_path, recursive = TRUE)
}
# 3. Set as working directory
setwd(subdir_path)
```

# t-test
```{r}
# 1. Log2 transform the data
#dma_data <- log2(processed_data)
dma_data <- processed_data

# 2. Match sample conditions
Conditions <- SSF$Conditions[match(rownames(dma_data), rownames(SSF))]

# 4. Filter sample indices
idx1 <- which(Conditions == group1)
idx2 <- which(Conditions == group2)

# 5. Perform t-test for each metabolite
results <- apply(dma_data, 2, function(x) {
  x1 <- x[idx1]
  x2 <- x[idx2]
  
  if(CoRe==T){
    mean1 <- mean(x1, na.rm = TRUE)
    mean2 <- mean(x2, na.rm = TRUE)
    
    NA0 <- any(x1 == 0, na.rm = TRUE) | any(x2 == 0, na.rm = TRUE)
    
    if(abs(mean1 - mean2)<=1){
      log2dist <- 0.1
    }else{
      log2dist <- log2(abs(mean1 - mean2))
    }

    if (abs(mean1) > abs(mean2)) log2dist <- log2dist * 1
    if (abs(mean1) < abs(mean2)) log2dist <- log2dist * -1
    
    if (mean1 > 0 & mean2 > 0) core_spec <- "Released"
    else if (mean1 < 0 & mean2 < 0) core_spec <- "Consumed"
    else if (mean1 > 0 & mean2 < 0) core_spec <- paste("Released in", group1, "and Consumed", group2)
    else if (mean1 < 0 & mean2 > 0) core_spec <- paste("Consumed in", group1, "and Released", group2)
    else core_spec <- "No Change"
    
    core_class <- ifelse(core_spec == "Released", "Released",
                         ifelse(core_spec == "Consumed", "Consumed", "Released/Consumed"))
    
    pval <- tryCatch(t.test(x1, x2)$p.value, error = function(e) NA)
    
    return(c(Log2_Distance = log2dist,
             p.val = pval,
             CoRe_specific = core_spec,
             CoRe = core_class,
             Mean_group1 = mean1,
             Mean_group2 = mean2))
  }else{
    pval <- tryCatch(t.test(x1, x2)$p.value, error = function(e) NA)
    log2FC <- mean(x1, na.rm = TRUE) - mean(x2, na.rm = TRUE)
    c(log2FC = log2FC, p.val = pval)
  }
  
})

# 6. Build final dataframe
DMA_results_df <- as.data.frame(t(results))
DMA_results_df$feature <- rownames(DMA_results_df)
DMA_results_df$p.adjusted <- p.adjust(DMA_results_df$p.val, method = "fdr")

# 7. Prepare output and save
if(CoRe==T){
  DMA_results_df <- DMA_results_df[, c("feature", "Log2_Distance", "p.val", "p.adjusted", "CoRe_specific", "CoRe",          "Mean_group1" ,  "Mean_group2" )]
  colnames(DMA_results_df)[1] <- "Metabolite"
  DMA_results_df$Log2_Distance <- as.numeric(DMA_results_df$Log2_Distance)
  #DMA_results_df[] <- lapply(DMA_results_df, function(x) if(is.numeric(x)) round(x, 4) else x)
  #
 # DMA_results_df[sapply(DMA_results_df, is.numeric)] <- round(DMA_results_df[sapply(DMA_results_df, is.numeric)], 8)
  DMA_results_df$Log2_Distance <- round(DMA_results_df$Log2_Distance, 3)
  write.csv(DMA_results_df, file = file.path(subdir_path, paste0("CoRe_t.test_", comp_name, ".csv")), row.names = FALSE)
  
}else{
  DMA_results_df <- DMA_results_df[, c("feature", "log2FC", "p.val", "p.adjusted")]
  colnames(DMA_results_df)[1] <- "Metabolite"
  #DMA_results_df[] <- lapply(DMA_results_df, function(x) if(is.numeric(x)) round(x, 4) else x)
  #DMA_results_df[sapply(DMA_results_df, is.numeric)] <- round(DMA_results_df[sapply(DMA_results_df, is.numeric)], 8)
  
  # num_mat <- as.matrix(DMA_results_df[sapply(DMA_results_df, is.numeric)])
  # min_val <- min(num_mat[num_mat > 0], na.rm = TRUE)
  # dec <- ceiling(-log10(min_val))
  # DMA_results_df[sapply(DMA_results_df, is.numeric)] <- round(DMA_results_df[sapply(DMA_results_df, is.numeric)], dec)


  DMA_results_df$log2FC <- round(DMA_results_df$log2FC, 3)
  write.csv(DMA_results_df, file = file.path(paste0(subdir_path, "/t.test_",comp_name,".csv")), row.names = FALSE)
}


DMA_results_df$feature <- DMA_results_df$Metabolite
DMA_results_df$p.val <- as.numeric(DMA_results_df$p.val)

```

# p Cut off selection
```{r}
DMA_results_df$neglog10_pval <- -log10(DMA_results_df$p.adjusted)

fc_cutoff <- 0.5
pval_cutoff =0.05


p.cutoff_selection <- ggplot(DMA_results_df, aes(x = reorder(feature, neglog10_pval), y = neglog10_pval)) +
  geom_col(fill = "darkred") +
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01), -log10(0.001), -log10(pval_cutoff)), 
             linetype = "dashed", color = "blue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "blue", size = 1) +  # Quadratic fit
  annotate("text", x = Inf, y = -log10(0.05), label = "p = 0.05", hjust = 12, vjust = -0.3, color = "blue") +
  annotate("text", x = Inf, y = -log10(0.01), label = "p = 0.01", hjust = 12, vjust = -0.3, color = "blue") +
  annotate("text", x = Inf, y = -log10(0.001), label = "p = 0.001", hjust = 10.4, vjust = -0.3, color = "blue") +
  annotate("text", x = Inf, y = -log10(pval_cutoff), label = paste0("p = ", pval_cutoff), hjust = 14.1, vjust = -0.3, color = "red") +
  labs(title = "-log10(p-value) by Feature", x = "Metabolite", y = "-log10(p-value)") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank()
  )

p.cutoff_selection

ggsave(paste0(subdir_path, "/p.adj_cutoff_selection.svg"), plot = p.cutoff_selection, width = 10, height = 8)


# 
# # Order data by p-value
# DMA_results_df <- DMA_results_df[order(DMA_results_df$neglog10_pval), ]
# DMA_results_df$feature <- factor(DMA_results_df$feature, levels = DMA_results_df$feature)
# DMA_results_df$feature_index <- seq_len(nrow(DMA_results_df))  # numeric for fitting
# 
# ggplot(DMA_results_df, aes(x = feature, y = neglog10_pval)) +
#   geom_col(fill = "darkred") +
#   geom_line(aes(x = feature, y = predict(lm(neglog10_pval ~ poly(feature_index, 2), data = DMA_results_df))),
#             color = "black", size = 1, group = 1) +
#   geom_hline(yintercept = c(-log10(0.001), -log10(0.01), -log10(0.05), -log10(x)),
#              linetype = "dashed", color = "blue") +
#   annotate("text", x = 1, y = -log10(0.001), label = "p = 0.001", hjust = -0.2, color = "blue") +
#   annotate("text", x = 1, y = -log10(0.01),  label = "p = 0.01",  hjust = -0.2, color = "blue") +
#   annotate("text", x = 1, y = -log10(0.05),  label = "p = 0.05",  hjust = -0.2, color = "blue") +
#   annotate("text", x = 1, y = -log10(x),     label = "p = 0.1",   hjust = -0.2, color = "red") +
#   labs(title = "-log10(p-value) by Feature with Quadratic Fit",
#        x = "Metabolite", y = "-log10(p-value)") +
#   theme(
#     axis.text.x = element_blank(),
#     axis.ticks.x = element_blank(),
#     panel.grid.major.x = element_blank()
#   )



```

# Volcanos 
```{r}
# Assuming your results are in a data.frame like this:
# DMA_results_df: columns = feature, log2FC, p.val, p.adjusted

if(CoRe==T){
  
  DMA_results_df$Significance <- ifelse(DMA_results_df$p.adjusted < pval_cutoff &
                                          abs(DMA_results_df$Log2_Distance) > fc_cutoff,
                                        ifelse(DMA_results_df$Log2_Distance > 0, "Up", "Down"),
                                        "NS")
  
  #DMA CORE PLOT
  p2 <-ggplot(DMA_results_df, aes(x = Log2_Distance, y = neglog10_pval, color = CoRe)) +
    geom_point(size = 2) +
    geom_text_repel(
      data = subset(DMA_results_df, Significance != "NS"),
      aes(label = feature),
      size = 3,
      max.overlaps = 50,
      box.padding = 0.3,
      segment.color = "grey50"
    ) +
    geom_hline(yintercept = -log10(pval_cutoff), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-fc_cutoff, fc_cutoff), linetype = "dashed", color = "black") +
    scale_color_manual(values = c(
      "Released" = "blue",
      "Consumed" = "green",
      "Released/Consumed" = "purple",
      "No Change" = "grey"
    )) +
    labs(
      title = paste(group1, "vs", group2),
      x = "Log2 Distance",
      y = "-Log10 Adjusted p-value"
    ) +
    theme_minimal(base_size = 14)
  
  
  ggsave(file.path(subdir_path, paste0("Volcano_CoRe_", comp_name, ".svg")), plot = p2, width = 10, height = 8)
  
  # DMA CORE ZOOMED PLOT
  p2_zoom <- ggplot(DMA_results_df, aes(x = Log2_Distance, y = neglog10_pval, color = CoRe)) +
    geom_point(size = 2) +
    geom_text_repel(
      data = subset(DMA_results_df, Significance != "NS"),
      aes(label = feature),
      size = 3,
      max.overlaps = 50,
      box.padding = 0.3,
      segment.color = "grey50"
    ) +
    geom_hline(yintercept = -log10(pval_cutoff), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-fc_cutoff, fc_cutoff), linetype = "dashed", color = "black") +
    scale_color_manual(values = c(
      "Released" = "blue",
      "Consumed" = "green",
      "Released/Consumed" = "purple",
      "No Change" = "grey"
    )) +
    labs(
      title = paste(group1, "vs", group2),
      x = "Log2 Distance",
      y = "-Log10 Adjusted p-value"
    ) +
    theme_minimal(base_size = 14)
  
  ggsave(file.path(subdir_path, paste0("Volcano_CoRe_", comp_name, "_ZOOM.svg")),
         plot = p2_zoom,  width = 28, height = 18, limitsize = FALSE)
  
  
  
  
  #DMA CORE PLOT
  p1 <-ggplot(DMA_results_df, aes(x = Log2_Distance, y = neglog10_pval)) +
    geom_point(aes(color = Significance), size = 2) +
    geom_text_repel(
      data = subset(DMA_results_df, Significance != "NS"),
      aes(label = feature),
      size = 3,
      max.overlaps = 50,
      box.padding = 0.3,
      segment.color = "grey50"
    ) +
    geom_hline(yintercept = -log10(pval_cutoff), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-fc_cutoff, fc_cutoff), linetype = "dashed", color = "black") +
    scale_color_manual(values = c("Up" = "red", "Down" = "blue", "NS" = "grey")) +
    labs(
      title = paste(group1, "vs", group2),
      x = "Log2 Distance",
      y = "-Log10 Adjusted p-value",
      color = "Regulation"
    ) +
    theme_minimal(base_size = 14)
  
  
  ggsave(file.path(subdir_path, paste0("Volcano_", comp_name, ".svg")), plot = p1,width = 10, height = 8)
  
  # DMA CORE ZOOMED PLOT
  p1_zoom <- ggplot(DMA_results_df, aes(x = Log2_Distance, y = neglog10_pval)) +
    geom_point(aes(color = Significance), size = 2) +
    geom_text_repel(
      data = subset(DMA_results_df, Significance != "NS"),
      aes(label = feature),
      size = 3,
      max.overlaps = 50,
      box.padding = 0.3,
      segment.color = "grey50"
    ) +
    geom_hline(yintercept = -log10(pval_cutoff), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-fc_cutoff, fc_cutoff), linetype = "dashed", color = "black") +
    scale_color_manual(values = c("Up" = "red", "Down" = "blue", "NS" = "grey")) +
    labs(
      title = paste(group1, "vs", group2),
      x = "Log2 Distance",
      y = "-Log10 Adjusted p-value",
      color = "Regulation"
    ) +
    theme_minimal(base_size = 14)
  
  ggsave(file.path(subdir_path, paste0("Volcano_", comp_name, "_ZOOM.svg")),
         plot = p1_zoom,  width = 28, height = 18, limitsize = FALSE)
  
  
  
  ###### note core
}else{
  VolcanoPlot <- EnhancedVolcano(
    DMA_results_df,
    lab = DMA_results_df$feature,
    x = 'log2FC',
    y = 'p.adjusted',
    title = paste0(gsub("_", " ", comp_name)),
    subtitle = 'Differential Metabolite Analysis (t-test)',
    xlab = bquote(~Log[2]~Fold~Change),
    ylab = bquote(~-Log[10]~Adjusted~p~value),
    pCutoff = pval_cutoff,
    FCcutoff = fc_cutoff,
    pointSize = 2.5,
    labSize = 3,
    colAlpha = 1,
    legendPosition = 'right',
    drawConnectors = T,
    ylim = c(0, max_y <- ceiling(max(-log10(DMA_results_df$p.adjusted), na.rm = TRUE) + 0.5))
  )
  
  
  ggsave(paste0(subdir_path, "/Volcano_",comp_name,".svg"), plot = VolcanoPlot, width = 10, height = 8)
  
  
  
  ##ZOOMED volcano
  VolcanoPlot <- EnhancedVolcano(
    DMA_results_df,
    lab = DMA_results_df$feature,
    x = 'log2FC',
    y = 'p.adjusted',
    title = paste0(gsub("_", " ", comp_name)),
    subtitle = 'Differential Metabolite Analysis (t-test)',
    xlab = bquote(~Log[2]~Fold~Change),
    ylab = bquote(~-Log[10]~Adjusted~p~value),
    pCutoff = pval_cutoff,
    FCcutoff = fc_cutoff,
    pointSize = 2.1,
    labSize = 5,
    colAlpha = 1,
    legendPosition = 'right',
    arrowheads = FALSE,
    drawConnectors = T,
    ylim = c(0, max_y <- ceiling(max(-log10(DMA_results_df$p.adjusted), na.rm = TRUE) + 0.5))
  )
  # Save plot
  ggsave(paste0(subdir_path, "/Volcano_",comp_name,"_ZOOM.svg"), plot = VolcanoPlot, width = 50, height = 32,limitsize = FALSE)
  
}


```

#Pathview
```{r}
if(CoRe==FALSE){
  Pathview_dir <- file.path(output.dir, "6. Pathview")
if (!dir.exists(Pathview_dir)) dir.create(Pathview_dir, recursive = TRUE)

subdir_path <- file.path(Pathview_dir, comp_name)
# 2. Create the subdirectory if it doesn't exist
if (!dir.exists(subdir_path)) {
  dir.create(subdir_path, recursive = TRUE)
}
# 3. Set as working directory
setwd(subdir_path)

rownames(universe) <- universe$Input.name
tmp <- merge(DMA_results_df,    universe%>% select(KEGG_ID),by=0)
fc_vector <- setNames(tmp$log2FC, tmp$KEGG_ID)
# Clean up NA values in your fc_vector
fc_vector <- fc_vector[!is.na(fc_vector)]
# List of KEGG pathway IDs
kegg_paths <- c(
  "hsa00010", "hsa00030", "hsa00020", "hsa00190",
  "hsa00250", "hsa00330", "hsa00260", "hsa00280",
  "hsa00230", "hsa00240", "hsa00480", "hsa00760", "hsa00520"
)

path_names <- c(
  hsa00010 = "Glycolysis",
  hsa00030 = "Pentose_Phosphate",
  hsa00020 = "TCA",
  hsa00190 = "OxPhos",
  hsa00250 = "Ala_Asp_Glu",
  hsa00330 = "Arg_Pro",
  hsa00260 = "Gly_Ser_Thr",
  hsa00280 = "Val_Leu_Ile",
  hsa00230 = "Purine",
  hsa00240 = "Pyrimidine",
  hsa00480 = "Glutathione",
  hsa00760 = "Nicotinate",
  hsa00520 = "Amino_Sugar"
)


for (path_id in names(path_names)) {
  pathview(
    cpd.data = fc_vector,
    pathway.id = path_id,
    species = "hsa",
    kegg.native = TRUE,
    out.suffix = path_names[[path_id]],
    cpd.idtype = "kegg"
  )
}

## remove unsessesary files
files <- list.files()
# Pattern: ends with number(s), dot, and exactly 3 lowercase letters
pattern <- "\\d+\\.[a-z]{3}$"
# Find matching files
files_to_remove <- files[grepl(pattern, files)]
# Remove them
file.remove(files_to_remove)

}
```

# Ora
```{r}
Ora_dir <- file.path(output.dir, "5. ORA")
if (!dir.exists(Ora_dir)) dir.create(Ora_dir, recursive = TRUE)

rm(mSet)
subdir_path <- file.path(Ora_dir, comp_name)
# 2. Create the subdirectory if it doesn't exist
if (!dir.exists(subdir_path)) {
  dir.create(subdir_path, recursive = TRUE)
}
# 3. Set as working directory
setwd(subdir_path)

# Make a list of metabs
if(CoRe==T){
  ora_list <- DMA_results_df %>% filter(abs(Log2_Distance)>0.5 & p.adjusted<pval_cutoff) %>% select(feature)  %>% unlist() %>% as.vector()
}else{
  ora_list <- DMA_results_df %>% filter(abs(log2FC)>0.5 & p.adjusted<pval_cutoff) %>% select(feature)  %>% unlist() %>% as.vector()
}
# Remove any asterisks
ora_list <- gsub("\\*", "", ora_list)
ora_list <- gsub("†", "", ora_list)

ora_list <- gsub(  "CAR 3:0 Propionylcarnitine", "Propionylcarnitine", ora_list)
ora_list <- gsub( "CAR 6:0 Hexanoylcarnitine" , "Hexanoylcarnitine", ora_list)

ora_list <- ora_list %>% na.omit()

#ora_list <- gsub("-", " ", ora_list)
checkdf <- data.frame(ora_list)

name <- comp_name

# Create vector consisting of compounds for enrichment analysis 
tmp.vec <- ora_list

# Create mSetObj
mSet<-InitDataObjects("conc", "msetora", FALSE, default.dpi = 300)

#Set up mSetObj with the list of compounds
mSet<-Setup.MapData(mSet, tmp.vec);

# Cross reference list of compounds against libraries (hmdb, pubchem, chebi, kegg, metlin)
mSet<-CrossReferencing(mSet, "name");

# Create the mapping results table
mSet<-CreateMappingResultTable(mSet)

Check <- read.csv(file.path(subdir_path, "name_map.csv"))
if(length(Check$Match[!is.na(Check$Match)])>5){
  # Set the metabolite filter
  mSet<-SetMetabolomeFilter(mSet, F);
  
  # # Select metabolite set library, refer to 
  # mSet<-SetCurrentMsetLib(mSet, "smpdb_pathway", 0); #kegg_pathway or smpdb_pathway
  # # Calculate hypergeometric score, results table generated in your working directory
  # mSet<-CalculateHyperScore(mSet)
  # or this
  # Initialize mSetObj for offline mode
  mSet <- InitDataObjects("conc", "msetora", FALSE, default.dpi = 300)  # Set TRUE for offline mode
  
  # Set pathway library location
  mSet <- SetCurrentMsetLib(mSet, "smpdb_pathway", 0)  # Replace "kegg_pathway" as needed
  
  # Load your pathway data locally
  # Ensure pathway files are downloaded and accessible in the working directory
  mSet <- CalculateHyperScore(mSet)
  
  ora_mat <-as.data.frame( mSet$analSet$ora.mat)
  # Add enrichment ratio column
  ora_mat$EnrichmentRatio <- ora_mat$hits / ora_mat$expected
  
  mSet$analSet$ora.mat <- ora_mat
  
  # Order by enrichment ratio (descending)
  mSet$analSet$ora.mat <- mSet$analSet$ora.mat[order(mSet$analSet$ora.mat$EnrichmentRatio, decreasing = TRUE), ]
  
  
  # Plot the ORA, bar-graph
  mSet<-PlotORA(mSet, paste0("Plot_ORA_",comp_name,"_smpdb_pathway_"), "bar", "pdf", 72, width=NA)
  
  OraRes <- mSet$analSet$ora.mat %>% as.data.frame()
  # OraRes <- rownames_to_column(OraRes, "Pathway")
  # writexl::write_xlsx(OraRes, paste0(name,"_OraRes.xlsx"))
  rm(mSet)
  
  # find metabs to select
  dt_saved <- qs::qread("smpdb_pathway.qs")
  writexl::write_xlsx(dt_saved, paste0("Path-Metab_assignment",comp_name,".xlsx"))
  
  setwd(ProjectFolder)
}
```



##### TST METAPROVIZ
```{r}
library(MetaProViz)
# MappingInfo <- MetaProViz::ToyData(Data="Cells_MetaData")
DMA_Annova <-  MetaProViz::dma(data=processed_data,
                               metadata_sample=SSF,
                               metadata_info = c(Conditions="Conditions", Numerator=NULL, Denominator = "Vehicle"),
                               pval ="aov",
                               padj="fdr",
                              # metadata_feature = MappingInfo,
                               core=TRUE)



# volcano with names
DMA_Vehicle_vs_Type_1 <-  DMA_Annova[["dma"]][["Vehicle_vs_Type_1"]]
# Run with default parameter --> only need to provide Input_data and the title we like
MetaProViz::viz_volcano(data=DMA_Vehicle_vs_Type_1%>%column_to_rownames("Metabolite"),
                       x= "Log2(Distance)", select_label = NULL)

```


### ----------- THE END -----------










# DMA correlation and lolipop
```{r}
DMA_dir <- file.path(output.dir, "4. DMA")
if (!dir.exists(DMA_dir)) dir.create(DMA_dir, recursive = TRUE)
# Pick your two files manually
csv_files <- list.files(path = DMA_dir,  pattern = "\\.csv$", full.names = TRUE, recursive = T)
# Select input files
# csv_files[1]
file1 <- csv_files[1]  # or replace with the actual index/name
# csv_files[2]
file2 <- csv_files[2]

# Load data
df1 <- read.csv(file1)
df2 <- read.csv(file2)

df1_sub <- df1[, c("Metabolite", "Log2_Distance")]
df2_sub <- df2[, c("Metabolite", "Log2_Distance")]

colnames(df1_sub) <- c("Metabolite", "log2FC_df1")
colnames(df2_sub) <- c("Metabolite", "log2FC_df2")


merged_logFC <- merge(df1_sub, df2_sub, by = "Metabolite")

merged_logFC <- merged_logFC %>%
  mutate(
    direction_df1 = sign(log2FC_df1),
    direction_df2 = sign(log2FC_df2),
    same_direction = direction_df1 == direction_df2,
    df2_higher_magnitude = abs(log2FC_df2) > abs(log2FC_df1),
    highlight = same_direction & df2_higher_magnitude
  )


# DMA_corr_plot <- ggplot(merged_logFC, aes(x = log2FC_df1, y = log2FC_df2)) +
#   geom_point(aes(color = highlight), size = 2) +
#   geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
#   labs(
#     x = paste0("Log2FC_" ,sub(".*t_test_(.*?)\\.csv$", "\\1", file1)),
#     y = paste0("Log2FC_" ,sub(".*t_test_(.*?)\\.csv$", "\\1", file2)),
#     title = "Comparison of log2 Fold Changes",
#     subtitle = paste0("Highlight: Same direction, higher magnitude in: " ,sub(".*t_test_(.*?)\\.csv$", "\\1", file2))
#   ) +
#   scale_color_manual(values = c("grey70", "dodgerblue")) +
#   coord_fixed(xlim = c(-8, 8), ylim = c(-10, 11)) +
#   theme_minimal() + theme(legend.position = "none")
# 
# ggsave(file.path(output.dir, "DMA_correlation_plot.svg"),  plot = DMA_corr_plot, width = 60, height = 40, units = "cm")

colnames(merged_logFC)[2] <- paste0(colnames(merged_logFC)[2],"_" ,sub( ".*t\\.test_([^/]+)\\.csv$", "\\1", file1) )
colnames(merged_logFC)[3] <- paste0(colnames(merged_logFC)[3],"_" ,sub( ".*t\\.test_([^/]+)\\.csv$", "\\1", file2) )
colnames(merged_logFC)[7] <- gsub("df2_", paste0(sub(".*t\\.test_([^/]+)\\.csv$", "\\1", file2), "_") ,colnames(merged_logFC)[7]  )
merged_logFC <- merged_logFC[,-c(4,5,8)]


SameDirection <- merged_logFC %>% filter(same_direction == TRUE)
OppositeDirection <- merged_logFC %>% filter(same_direction != TRUE)

write.csv(merged_logFC, file.path(output.dir, "DMA_correlations_log2FC.csv"), row.names = FALSE)

```

# lolipop
```{r}
data <- SameDirection # or OppositeDirection
# Reshape to long format
long_df <- data %>%
  pivot_longer(
    cols = starts_with("log2FC_"),
    names_to = "Condition",
    values_to = "log2FC"
  )

# Optional: Clean up condition labels
long_df$Condition <- gsub("log2FC_df1_","", long_df$Condition)
long_df$Condition <- gsub("log2FC_df2_", "",long_df$Condition)

# Sort metabolites by the Homozygous condition log2FC values
sorted_metabolites <- data %>%
  arrange(desc(colnames(data)[2])) %>%
  pull(Metabolite)

# Reorder the factor levels
long_df$Metabolite <- factor(long_df$Metabolite, levels = sorted_metabolites)

# Plot
lolipop <- ggplot(long_df, aes(x = log2FC, y = Metabolite, color = Condition)) +
  geom_segment(aes(x = 0, xend = log2FC, y = Metabolite, yend = Metabolite),
               size = 1.2) +
  geom_point(size = 3) +
  labs(
    x = "log2 Distance",
    y = "Metabolite",
    title = "Lollipop Plot of Same direction log2 Fold Changes",
    color = "Condition"
  ) +
  theme_minimal()

ggsave(file.path(output.dir, "DMA_lolipop_same_direction.svg"),  plot = lolipop, width = 30, height = 40, units = "cm")


data <- OppositeDirection # or OppositeDirection
# Reshape to long format
long_df <- data %>%
  pivot_longer(
    cols = starts_with("log2FC_"),
    names_to = "Condition",
    values_to = "log2FC"
  )

# Optional: Clean up condition labels
long_df$Condition <- gsub("log2FC_df1_","", long_df$Condition)
long_df$Condition <- gsub("log2FC_df2_", "",long_df$Condition)

# Sort metabolites by the Homozygous condition log2FC values
sorted_metabolites <- data %>%
  arrange(desc(colnames(data)[2])) %>%
  pull(Metabolite)

# Reorder the factor levels
long_df$Metabolite <- factor(long_df$Metabolite, levels = sorted_metabolites)

# Plot
lolipop <- ggplot(long_df, aes(x = log2FC, y = Metabolite, color = Condition)) +
  geom_segment(aes(x = 0, xend = log2FC, y = Metabolite, yend = Metabolite),
               size = 1.2) +
  geom_point(size = 3) +
  labs(
    x = "log2 Distancee",
    y = "Metabolite",
    title = "Lollipop Plot of Same direction log2 Fold Changes",
    color = "Condition"
  ) +
  theme_minimal()

ggsave(file.path(output.dir, "DMA_lolipop_Opposite_direction.svg"),  plot = lolipop, width = 30, height = 40, units = "cm")
```

# mixed models test
```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(lme4)
  library(lmerTest)
  library(ggplot2)
  library(ggrepel)
})

# =========================
# INPUT OBJECTS YOU HAVE:
#   SSFsave   (metadata; rownames are sample IDs)
#   data_norm (matrix/data.frame; rownames are sample IDs)
# =========================

# ---- settings ----
fc_cutoff  <- 0.5
fdr_cutoff <- 0.1

output.dir <- output.dir  # you already have this in your pipeline
DMA_dir <- file.path(output.dir, "4. DMA_MixedModel_Unmerged")
dir.create(DMA_dir, recursive = TRUE, showWarnings = FALSE)

# ---- align sample info to data ----
stopifnot(all(rownames(data_norm) %in% rownames(SSFsave)))

sample_info <- SSFsave[match(rownames(data_norm), rownames(SSFsave)), , drop=FALSE]
sample_info$sample_id <- rownames(data_norm)

sample_info$Conditions <- factor(sample_info$Conditions)
sample_info$`Biological Replicate` <- factor(sample_info$`Biological Replicate`)
sample_info$`Technical Replicate`  <- factor(sample_info$`Technical Replicate`)

# OPTIONAL: if you want log2 scale, uncomment:
# data_norm <- log2(data_norm)

# =========================
# function: one contrast
# =========================
run_lmer_contrast <- function(mat, sample_info,
                              group1, group2,
                              outdir,
                              fc_cutoff = 0.5,
                              fdr_cutoff = 0.1) {
  
  dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
  
  # subset samples for the two groups
  keep <- sample_info$Conditions %in% c(group1, group2)
  si   <- droplevels(sample_info[keep, ])
  
  # set ref so coefficient = group1 - group2
  si$Conditions <- relevel(factor(si$Conditions), ref = group2)
  
  mat_sub <- mat[si$sample_id, , drop = FALSE]
  
  # long format
  df_long <- mat_sub |>
    as.data.frame() |>
    tibble::rownames_to_column("sample_id") |>
    pivot_longer(-sample_id, names_to = "Metabolite", values_to = "value") |>
    left_join(
      si |> select(sample_id, Conditions, `Biological Replicate`, `Technical Replicate`),
      by = "sample_id"
    )
  
  # fit per metabolite
  res <- df_long |>
    group_by(Metabolite) |>
    group_modify(~{
      d <- .x
      if (all(is.na(d$value))) return(tibble(log2FC = NA_real_, p.val = NA_real_))
      
      # mixed model: tech reps are repeated measures within the same bio replicate
      fit <- tryCatch(
        lmer(value ~ Conditions + (1 | `Biological Replicate`), data = d, REML = FALSE),
        error = function(e) NULL
      )
      
      if (is.null(fit)) {
        # fallback (rare): use t-test on group means per biological replicate
        d2 <- d |>
          group_by(Conditions, `Biological Replicate`) |>
          summarise(v = mean(value, na.rm=TRUE), .groups="drop")
        
        x1 <- d2$v[d2$Conditions == group1]
        x2 <- d2$v[d2$Conditions == group2]
        
        est  <- mean(x1, na.rm=TRUE) - mean(x2, na.rm=TRUE)
        pval <- tryCatch(t.test(x1, x2)$p.value, error=function(e) NA_real_)
        return(tibble(log2FC = est, p.val = pval))
      }
      
      coefs <- summary(fit)$coefficients
      term  <- paste0("Conditions", group1)
      
      if (!term %in% rownames(coefs)) return(tibble(log2FC = NA_real_, p.val = NA_real_))
      
      tibble(
        log2FC = unname(coefs[term, "Estimate"]),
        p.val  = unname(coefs[term, "Pr(>|t|)"])
      )
    }) |>
    ungroup() |>
    mutate(
      p.adjusted   = p.adjust(p.val, method = "fdr"),
      neglog10_fdr = -log10(p.adjusted),
      Comparison   = paste0(group1, "_Vs_", group2),
      Significance = case_when(
        !is.na(p.adjusted) & p.adjusted < fdr_cutoff & abs(log2FC) > fc_cutoff & log2FC > 0 ~ "Up",
        !is.na(p.adjusted) & p.adjusted < fdr_cutoff & abs(log2FC) > fc_cutoff & log2FC < 0 ~ "Down",
        TRUE ~ "NS"
      )
    )
  
  # save csv
  out_csv <- file.path(outdir, paste0("lmer_", group1, "_Vs_", group2, ".csv"))
  write.csv(res, out_csv, row.names = FALSE)
  
  # volcano
  p <- ggplot(res, aes(x = log2FC, y = neglog10_fdr)) +
    geom_point(aes(color = Significance), size = 2) +
    geom_hline(yintercept = -log10(fdr_cutoff), linetype = "dashed") +
    geom_vline(xintercept = c(-fc_cutoff, fc_cutoff), linetype = "dashed") +
    scale_color_manual(values = c(Up="red", Down="blue", NS="grey")) +
    labs(
      title = paste0(group1, " vs ", group2),
      subtitle = "lmer: value ~ Condition + (1|Biological Replicate)",
      x = "Effect (group1 - group2)",
      y = "-log10(FDR)"
    ) +
    theme_minimal(base_size = 14)
  
  p_lab <- p +
    geom_text_repel(
      data = res |> filter(Significance != "NS"),
      aes(label = Metabolite),
      size = 3,
      max.overlaps = 50,
      box.padding = 0.3,
      segment.color = "grey50"
    )
  
  ggsave(file.path(outdir, paste0("Volcano_lmer_", group1, "_Vs_", group2, ".svg")),
         plot = p_lab, width = 10, height = 8)
  
  ggsave(file.path(outdir, paste0("Volcano_lmer_", group1, "_Vs_", group2, "_ZOOM.svg")),
         plot = p_lab, width = 28, height = 18, limitsize = FALSE)
  
  invisible(res)
}

# =========================
# RUN BOTH COMPARISONS
# =========================
res_Type1 <- run_lmer_contrast(
  mat = data_norm,
  sample_info = sample_info,
  group1 = "Type_1",
  group2 = "Vehicle",
  outdir = file.path(DMA_dir, "Type_1_Vs_Vehicle"),
  fc_cutoff = fc_cutoff,
  fdr_cutoff = fdr_cutoff
)

res_Type2 <- run_lmer_contrast(
  mat = data_norm,
  sample_info = sample_info,
  group1 = "Type_2",
  group2 = "Vehicle",
  outdir = file.path(DMA_dir, "Type_2_Vs_Vehicle"),
  fc_cutoff = fc_cutoff,
  fdr_cutoff = fdr_cutoff
)


```



# Session information
```{r session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```

# Bibliography

